---
title: "figures_and_tables"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r install and load packages}

  if(!require(BiocManager)){install.packages("BiocManager")}
  if(!require(ggtree)){BiocManager::install("ggtree")}
  if(!require(DESeq2)){BiocManager::install("DESeq2")}
  library(ggtree)
  library(DESeq2)

  devtools::install_github("wilkox/ggwrap")
  library(ggwrap)
  devtools::install_github("clauswilke/ggtext")
  library(ggtext)
  remotes::install_github("larssnip/microclass")
  library(microclass)

  if (!require("pacman")){install.packages("pacman")}
  pacman::p_load(BiocGenerics, ComplexUpset, dendextend, devtools, flextable, furrr, GGally, ggdendro,  gggenes, ggplotify, ggpubr, grid, here, lemon, lme4, lmerTest, magrittr, MuMIn, pals, phangorn, phyloseq, RColorBrewer, readxl, rstatix, seqinr, sjPlot, spaa, cowplot, tidyverse, vegan, vroom, xlsx)

```

```{r Options}

  i_am("figures_and_tables.Rmd") # set where here() starts as the directory which contains Scripts/figures_and_tables.Rmd

  theme_set(theme_bw()) # set classic theme for ggplot
  browns_palette <- c("#D4B23C", "#D4A13C", "#D48D3C", "#B3621E", "#7A3500", "#130000")
  set.seed(123)

```

## Methods

# Supplementary Table S1 - Assembly statistics for sample libraries

```{r Supplementary Table S1 - Assembly statistics for sample libraries}

  assembly_statistics <- 
    as_tibble(read.csv(here("Data/assembly_statistics.csv")))

  write.csv(assembly_statistics, file = here("Tables/Table S1.csv"), row.names = F)

```

# Supplementary Table S2 - DNA vOTU raw reads

```{r Supplementary Table S2 - DNA vOTU raw reads}

  DNA_vOTU_reads <- 
    as_tibble(read.csv(here("Data/DNA_vOTU_reads.csv")))
  # 10,196 DNA vOTUs
    # 9,804 dsDNA phage vOTUs
    # 532 ssDNA vOTUs
  # raw genome read values (filtered for 1x coverage across 75% of genome sequence)

  write.csv(DNA_vOTU_reads, file = here("Tables/Table S2.csv"), row.names = F)

```

## Soil viral communities were structured with soil depth at both the population-level and strain-level

# Supplementary Figure S1 - Taxonomic novelty of recovered soil vOTUs

```{r Import and parse vConTACT2 network files}

# import nodes (17,917)
  nodes <- 
    as_tibble(read.delim(here("Data/nodes.csv"), sep = ",", header = T)) %>% 
    select(GENOME, x, y)

# import edges (1,721,928)
  edges <- 
    as_tibble(read.delim(here("Data/edges.csv"), sep = ",", header = T)) %>% 
    inner_join(nodes, by = "GENOME")

# record viral cluster statuses
  genome_by_genome <- 
    as_tibble(read.delim(here("Data/genome_by_genome_overview.csv"), sep = ",", header = T)) %>% 
    select("NAME" = Genome,
           "VC_STATUS" = VC.Status,
           VC) %>% 
    mutate(VC_STATUS = case_when(grepl("Overlap", VC_STATUS) ~ "Singleton",
                                 VC_STATUS == "Clustered/Singleton" ~ "Clustered",
                                 TRUE ~ VC_STATUS))
  # Clustered = high-confidence clustering (roughly equivalent to an ICTV genus).
  # Singleton = few or no gene similarities against other genomes (most singletons don’t make it into the network).
  # Overlap = sharing overlap with other genome(s) from multiple VCs. Often, these viruses have shared core genes, or a large portion of their genome has a conserved region that is shared among many others.
  # Outlier = some genes shared with other genomes, but ClusterONE wasn’t confident enough to place them within a particular VC (roughly related at the sub-family or family level).
  # Clustered/Singleton = in a subcluster alone (the cluster got split).
  
# create data frame of node location for genomes in viral clusters
  VC_nodes <-
    nodes %>%
    left_join(genome_by_genome, by = c("GENOME" = "NAME")) %>%
    mutate(REF_PHAGE = case_when(grepl("DNA_vOTU", GENOME) ~ "NO",
                                 TRUE ~ "YES")) %>% 
    mutate(PHAGE = case_when(GENOME %in% filter(DNA_vOTU_reads, PHAGE == "dsDNAphage")$NAME ~ "dsDNA vOTU",
                             GENOME %in% filter(DNA_vOTU_reads, PHAGE == "ssDNA")$NAME ~ "ssDNA vOTU",
                             TRUE ~ "Reference genome")) %>% 
    mutate(PHAGE = factor(PHAGE, levels = c("dsDNA vOTU", "ssDNA vOTU", "Reference genome")))
  
```

```{r Supplementary Figure S1A - Viral cluster identity}

# extract viral cluster identities for vOTUs
  cluster_identities <- 
    DNA_vOTU_reads %>% 
    group_by(CLUSTER) %>% 
    mutate(CLUSTER_STATUS = case_when(VC_STATUS == "Singleton" ~ "Singleton",
                                      VC_STATUS == "Outlier" & REF_PHAGES_IN_VC == "YES" ~ "Outlier to cluster with reference phages",
                                      VC_STATUS == "Outlier" & REF_PHAGES_IN_VC == "NO" ~ "Outlier to cluster without reference phages",
                                      VC_STATUS == "Clustered" & REF_PHAGES_IN_VC == "YES" ~ "Clustered with reference phages",
                                      VC_STATUS == "Clustered" & REF_PHAGES_IN_VC == "NO" ~ "Clustered without reference phages")) %>% 
    ungroup() %>% 
    select(NAME, CLUSTER_STATUS)

# create data frame with cluster status for visualising shared protein network
  df <- 
    VC_nodes %>% 
    left_join(cluster_identities, by = c("GENOME" = "NAME")) %>%
    mutate(CLUSTER_STATUS = case_when(is.na(CLUSTER_STATUS) ~ "INPHARED phage genome",
                                      CLUSTER_STATUS == "Clustered without reference phages" | CLUSTER_STATUS == "Outlier to cluster without reference phages" ~ "Without reference phages",
                                      CLUSTER_STATUS == "Clustered with reference phages" | CLUSTER_STATUS == "Outlier to cluster with reference phages" ~ "With reference phages",
                                      TRUE ~ CLUSTER_STATUS))

# plot shared protein network
  left <-
    ggplot(df, aes(x = x, y = y)) + 
      geom_line(data = edges, aes(group = PAIR),
                color = "black", linewidth = 0.5, alpha = 0.2) +
      geom_point(data = filter(df, CLUSTER_STATUS == "INPHARED phage genome"), aes(fill = CLUSTER_STATUS),
                 size = 2, shape = 21, stroke = 0.5, colour = "black", alpha = 0.8) +
      geom_point(data = filter(df, CLUSTER_STATUS == "Singleton"), aes(fill = CLUSTER_STATUS),
                 size = 2, shape = 21, stroke = 0.5, colour = "black", alpha = 0.8) +
      geom_point(data = filter(df, CLUSTER_STATUS == "Without reference phages"), aes(fill = CLUSTER_STATUS),
                 size = 2, shape = 21, stroke = 0.5, colour = "black", alpha = 0.8) +
      geom_point(data = filter(df, CLUSTER_STATUS == "With reference phages"), aes(fill = CLUSTER_STATUS),
                 size = 2, shape = 21, stroke = 0.5, colour = "black", alpha = 0.8) +
      scale_fill_manual(values = c("#009E73", "#D55E00", "#CC79A7", "grey80"),
                        limits = c("With reference phages", "Without reference phages", "Singleton", "INPHARED phage genome")) +
      theme_minimal() + 
      coord_fixed(xlim = c(0, 1),
                  ylim = c(0, 1)) +
      guides(fill = guide_legend(override.aes = list(size = c(4, 4, 4, 4)))) +
      theme(axis.text = element_blank(),
            axis.title = element_blank(),
            text = element_text(size = 10),
            panel.grid = element_blank(), 
            legend.direction = "vertical",
            legend.position = "right") +
      labs(fill = "Viral cluster indentity:")
  
  legend <- get_legend(left)
  left <- left + theme(legend.position = "none")
  
# plot bar chart of viral cluster identities for dsDNA and ssDNA vOTUs
  plot <- 
    DNA_vOTU_reads %>% 
    group_by(CLUSTER) %>% 
    mutate(CLUSTER_STATUS = case_when(VC_STATUS == "Singleton" ~ "Singleton",
                                      VC_STATUS == "Outlier" & REF_PHAGES_IN_VC == "YES" ~ "Clustered/outlier to cluster with reference phages",
                                      VC_STATUS == "Outlier" & REF_PHAGES_IN_VC == "NO" ~ "Clustered/outlier to cluster without reference phages",
                                      VC_STATUS == "Clustered" & REF_PHAGES_IN_VC == "YES" ~ "Clustered/outlier to cluster with reference phages",
                                      VC_STATUS == "Clustered" & REF_PHAGES_IN_VC == "NO" ~ "Clustered/outlier to cluster without reference phages")) %>% 
    ungroup() %>% 
    mutate(CLUSTER_STATUS = case_when(NAME %in% VC_nodes$GENOME ~ CLUSTER_STATUS,
                                      TRUE ~ "NONE")) %>%
    ggplot(aes(x = PHAGE, fill = factor(CLUSTER_STATUS, levels = c("NONE", "Clustered/outlier to cluster with reference phages", "Clustered/outlier to cluster without reference phages", "Singleton")))) +
      geom_bar(stat = "count", position = "fill") +
      scale_fill_manual(values = c(alpha("white", 0), "#009E73", "#D55E00", "#CC79A7"),
                        limits = c("NONE", "Clustered/outlier to cluster with reference phages", "Clustered/outlier to cluster without reference phages", "Singleton"),
                        labels = c("NONE", "With reference phages", "Without reference phages", "Singleton")) +
      scale_x_discrete(breaks = c("dsDNAphage", "ssDNA"),
                       labels = c("dsDNA", "ssDNA")) +
      labs(x = "vOTU group",
           y = "Proportion of vOTUs\nwithin group",
           fill = "Viral cluster identity:") +
      theme(legend.position = "none",
            text = element_text(size = 8))
  
  right <- plot_grid(legend, plot, ncol = 1, rel_heights = c(1, 1))
  
  A <- plot_grid(left, right, nrow = 1, rel_widths = c(2, 1), labels = "A", label_size = 15)

```

```{r Supplementary Figure S1B - Depth enrichment (Garry Oak)}

# extract depth enrichment in Garry Oak samples for vOTUs
  depth_enrichment <- 
    DNA_vOTU_reads %>% 
    mutate(JUMBO = case_when(LENGTH > 200000 ~ "YES",
                             TRUE ~ "NO")) %>% 
    select(NAME, GARRY_OAK_ENRICHMENT, JUMBO)

# create data frame with depth enrichment for visualising shared protein network
  df <- 
    VC_nodes %>% 
    left_join(depth_enrichment, by = c("GENOME" = "NAME")) %>% 
    filter(!is.na(GARRY_OAK_ENRICHMENT))

# plot shared protein network
  left <-
    ggplot(df, aes(x = x, y = y)) + 
      geom_line(data = filter(edges, GENOME %in% df$GENOME) %>% group_by(PAIR) %>% filter(n() > 1), aes(group = PAIR),
                color = "black", size = 0.5, alpha = 0.2) +
      geom_point(data = filter(df, GARRY_OAK_ENRICHMENT == "SUBSURFACE"), aes(fill = GARRY_OAK_ENRICHMENT, shape = JUMBO, size = JUMBO),
                 stroke = 0.5, colour = "black", alpha = 0.8) +
      geom_point(data = filter(df, GARRY_OAK_ENRICHMENT == "SURFACE"), aes(fill = GARRY_OAK_ENRICHMENT, shape = JUMBO, size = JUMBO),
                 stroke = 0.5, colour = "black", alpha = 0.8) +
      scale_fill_manual(values = c("#B5651D", "#2A0001"),
                        limits = c("SURFACE", "SUBSURFACE"),
                        labels = c("Surface-enriched", "Subsurface-enriched")) +
      scale_shape_manual(values = c(21, 23),
                         limits = c("NO", "YES"),
                         labels = c("Non-jumbo", "Jumbo phages"),
                         name = NULL) +
      scale_size_manual(values = c(2, 4),
                        limits = c("NO", "YES"),
                        labels = c("Non-jumbo", "Jumbo phages"),
                        name = NULL) +
      theme_minimal() + 
      coord_fixed(xlim = c(0, 1),
                  ylim = c(0, 1)) +
      guides(fill = guide_legend(override.aes = list(size = c(4, 4),
                                                     shape = c(21, 21)))) +
      theme(axis.text = element_blank(),
            axis.title = element_blank(),
            text = element_text(size = 10),
            panel.grid = element_blank(), 
            legend.direction = "vertical",
            legend.position = "right") +
      labs(fill = "Garry Oak enrichment:",
           shape = "",
           size = "")
  
  legend <- get_legend(left)
  left <- left + theme(legend.position = "none")
  
# plot bar chart of viral cluster identities for dsDNA and ssDNA vOTUs
  plot <-
    DNA_vOTU_reads %>% 
    mutate(GARRY_OAK_ENRICHMENT = case_when(is.na(GARRY_OAK_ENRICHMENT) ~ "NONE",
                                            TRUE ~ GARRY_OAK_ENRICHMENT)) %>% 
    ggplot(aes(x = PHAGE, fill = factor(GARRY_OAK_ENRICHMENT, levels = c("NONE", "SURFACE", "SUBSURFACE")))) +
      geom_bar(stat = "count", position = "fill") +
      scale_fill_manual(values = c(alpha("white", 0), "#B5651D", "#2A0001"),
                        limits = c("NONE", "SURFACE", "SUBSURFACE"),
                        labels = c(NA, "Surface-enriched", "Subsurface-enriched")) +
      scale_x_discrete(breaks = c("dsDNAphage", "ssDNA"),
                       labels = c("dsDNA", "ssDNA")) +
      labs(x = "vOTU group",
           y = "Proportion of vOTUs\nwithin group",
           fill = "Depth enrichment:") +
      theme(legend.position = "none",
            text = element_text(size = 8))
  
  right <- plot_grid(legend, plot, ncol = 1, rel_heights = c(1, 1))
  
  B <- plot_grid(left, right, nrow = 1, rel_widths = c(2, 1), labels = "B", label_size = 15)

```

```{r Supplementary Figure S1C - Depth enrichment (Hilly grassland)}

# extract depth enrichment in Hilly grassland samples for vOTUs
  depth_enrichment <- 
    DNA_vOTU_reads %>% 
    mutate(JUMBO = case_when(LENGTH > 200000 ~ "YES",
                             TRUE ~ "NO")) %>% 
    select(NAME, HILLY_GRASSLAND_ENRICHMENT, JUMBO)

# create data frame with depth enrichment for visualising shared protein network
  df <- 
    VC_nodes %>% 
    left_join(depth_enrichment, by = c("GENOME" = "NAME")) %>% 
    filter(!is.na(HILLY_GRASSLAND_ENRICHMENT))

# plot shared protein network
  left <-
    ggplot(df, aes(x = x, y = y)) + 
      geom_line(data = filter(edges, GENOME %in% df$GENOME) %>% group_by(PAIR) %>% filter(n() > 1), aes(group = PAIR),
                color = "black", size = 0.5, alpha = 0.2) +
      geom_point(data = filter(df, HILLY_GRASSLAND_ENRICHMENT == "SUBSURFACE" & JUMBO == "NO"), aes(fill = HILLY_GRASSLAND_ENRICHMENT, shape = JUMBO, size = JUMBO),
                 stroke = 0.5, colour = "black", alpha = 0.8) +
      geom_point(data = filter(df, HILLY_GRASSLAND_ENRICHMENT == "SURFACE" & JUMBO == "NO"), aes(fill = HILLY_GRASSLAND_ENRICHMENT, shape = JUMBO, size = JUMBO),
                 stroke = 0.5, colour = "black", alpha = 0.8) +
      geom_point(data = filter(df, HILLY_GRASSLAND_ENRICHMENT == "SUBSURFACE" & JUMBO == "YES"), aes(fill = HILLY_GRASSLAND_ENRICHMENT, shape = JUMBO, size = JUMBO),
                 stroke = 0.5, colour = "white", alpha = 0.8) +
      geom_point(data = filter(df, HILLY_GRASSLAND_ENRICHMENT == "SURFACE" & JUMBO == "YES"), aes(fill = HILLY_GRASSLAND_ENRICHMENT, shape = JUMBO, size = JUMBO),
                 stroke = 0.5, colour = "white", alpha = 0.8) +
      scale_fill_manual(values = c("#B5651D", "#2A0001"),
                        limits = c("SURFACE", "SUBSURFACE"),
                        labels = c("Surface-enriched", "Subsurface-enriched")) +
      scale_shape_manual(values = c(21, 23),
                         limits = c("NO", "YES"),
                         labels = c("Non-jumbo", "Jumbo phages"), 
                         name = NULL) +
      scale_size_manual(values = c(2, 4),
                        limits = c("NO", "YES"),
                        labels = c("Non-jumbo", "Jumbo phages"), 
                        name = NULL) +
      theme_minimal() + 
      coord_fixed(xlim = c(0, 1),
                  ylim = c(0, 1)) +
      guides(fill = guide_legend(override.aes = list(size = c(4, 4),
                                                     shape = c(21, 21),
                                                     colour = c("black", "black"),
                                                     alpha = c(0.8, 0.8))),
             shape = guide_legend(override.aes = list(size = c(2, 4),
                                                      stroke = c(0.5, 0.5),
                                                      colour = c("black", "black")))) +
      theme(axis.text = element_blank(),
            axis.title = element_blank(),
            text = element_text(size = 10),
            panel.grid = element_blank(), 
            legend.spacing.y = unit(1, "mm"),
            legend.direction = "vertical",
            legend.position = "right") +
      labs(fill = "Hilly grassland enrichment:")
  
  legend <- get_legend(left)
  left <- left + theme(legend.position = "none")
  
# plot bar chart of viral cluster identities for dsDNA and ssDNA vOTUs
  plot <-
    DNA_vOTU_reads %>% 
    mutate(HILLY_GRASSLAND_ENRICHMENT = case_when(is.na(HILLY_GRASSLAND_ENRICHMENT) ~ "NONE",
                                                  TRUE ~ HILLY_GRASSLAND_ENRICHMENT)) %>% 
    ggplot(aes(x = PHAGE, fill = factor(HILLY_GRASSLAND_ENRICHMENT, levels = c("NONE", "SURFACE", "SUBSURFACE")))) +
      geom_bar(stat = "count", position = "fill") +
      scale_fill_manual(values = c(alpha("white", 0), "#B5651D", "#2A0001"),
                        limits = c("NONE", "SURFACE", "SUBSURFACE"),
                        labels = c(NA, "Surface-enriched", "Subsurface-enriched")) +
      scale_x_discrete(breaks = c("dsDNAphage", "ssDNA"),
                       labels = c("dsDNA", "ssDNA")) +
      labs(x = "vOTU group",
           y = "Proportion of vOTUs\nwithin group",
           fill = "Depth enrichment:") +
      theme(legend.position = "none",
            text = element_text(size = 8))
  
  right <- plot_grid(legend, plot, ncol = 1, rel_heights = c(1, 1))
  
  C <- plot_grid(left, right, nrow = 1, rel_widths = c(2, 1), labels = "C", label_size = 15)

```

```{r Supplementary Figure S1D - vOTUs carrying AMGs}

# import viral gene annotations
  DNA_vOTU_gene_annotations <- 
    as_tibble(read.delim(file = here("Data/DNA_vOTU_gene_annotations.csv"), sep = ","))

# filter DRAMv annotations for AMGs
  AMG_annotations <-
    DNA_vOTU_gene_annotations %>% 
    left_join(select(DNA_vOTU_reads, c(NAME, LENGTH, CHECKV_QUALITY)), by = "NAME") %>% 
    filter(auxiliary_score == 1 | # at least one hallmark gene on the left and right flank
           auxiliary_score == 2 | # viral gene and viral-like gene on flanks
           auxiliary_score == 3) %>% # viral-like genes on both flanks
    filter(grepl("M", amg_flags)) %>% # keep hits involved in metabolism
    filter(grepl("F", amg_flags)) %>% # keep hits where the gene is within 5000 bases of the end of the contig
    filter(LENGTH >= 10000 | CHECKV_QUALITY == "Complete") %>% # likely complete viral genomes
    select(NAME, LOCUS, LENGTH, CHECKV_QUALITY, gene_position:PHROG_ANNOTATION)

  write.csv(AMG_annotations, file = here("Data/AMG_annotations.csv"), row.names = F)
  
# create data frame with AMG carriage for visualising shared protein network
  df <- 
    VC_nodes %>% 
    filter(GENOME %in% AMG_annotations$NAME) %>% 
    mutate(AMG = case_when(GENOME %in% filter(AMG_annotations, cazy_hits != "")$NAME ~ "CAZyme",
                           TRUE ~ "Non-CAZyme")) %>% 
    mutate(JUMBO = case_when(GENOME %in% filter(DNA_vOTU_reads, LENGTH > 200000)$NAME ~ "YES",
                             TRUE ~ "NO"))

# plot shared protein network
  left <-
    ggplot(df, aes(x = x, y = y)) + 
      geom_line(data = filter(edges, GENOME %in% df$GENOME) %>% group_by(PAIR) %>% filter(n() > 1), aes(group = PAIR),
                color = "black", size = 0.5, alpha = 0.2) +
      geom_point(data = filter(df, JUMBO == "NO"), aes(fill = AMG, shape = JUMBO, size = JUMBO),
                 stroke = 0.5, colour = "black", alpha = 0.8) + # fill = "#7570B3", 
      geom_point(data = filter(df, JUMBO == "YES"), aes(fill = AMG, shape = JUMBO, size = JUMBO),
                 stroke = 0.5, colour = "white", alpha = 0.8) + # fill = "#7570B3", 
      scale_fill_manual(values = c("#E7298A", "#66A61E"),
                        limits = c("CAZyme", "Non-CAZyme")) +
      scale_shape_manual(values = c(21, 23),
                         limits = c("NO", "YES"),
                         labels = c("Non-jumbo", "Jumbo phages"),
                         name = NULL) +
      scale_size_manual(values = c(2, 4),
                        limits = c("NO", "YES"),
                        labels = c("Non-jumbo", "Jumbo phages"),
                        name = NULL) +
      theme_minimal() + 
      coord_fixed(xlim = c(0, 1),
                  ylim = c(0, 1)) +
      guides(fill = guide_legend(override.aes = list(size = c(4, 4),
                                                     shape = c(21, 21),
                                                     colour = c("black", "black"),
                                                     alpha = c(0.8, 0.8))),
             shape = guide_legend(override.aes = list(size = c(2, 4),
                                                      stroke = c(0.5, 0.5),
                                                      colour = c("black", "black")))) +
      theme(axis.text = element_blank(),
            axis.title = element_blank(),
            text = element_text(size = 10),
            panel.grid = element_blank(), 
            legend.spacing.y = unit(1, "mm"),
            legend.direction = "vertical",
            legend.position = "right") +
      labs(fill = "AMG carriage:")
  
  legend <- get_legend(left)
  left <- left + theme(legend.position = "none")
  
# plot bar chart of viral cluster identities for dsDNA and ssDNA vOTUs
  plot <-
    DNA_vOTU_reads %>% 
    mutate(AMG = case_when(NAME %in% filter(AMG_annotations, cazy_hits != "")$NAME ~ "CAZyme",
                           NAME %in% filter(AMG_annotations, cazy_hits == "")$NAME ~ "Non-CAZyme",
                           TRUE ~ "No AMGs")) %>% 
    ggplot(aes(x = PHAGE, fill = factor(AMG, levels = c("No AMGs", "Non-CAZyme", "CAZyme")))) +
      geom_bar(stat = "count", position = "fill") +
      scale_fill_manual(values = c(alpha("white", 0), "#E7298A", "#66A61E"),
                        limits = c("No AMGs", "Non-CAZyme", "CAZyme")) +
      scale_x_discrete(breaks = c("dsDNAphage", "ssDNA"),
                       labels = c("dsDNA", "ssDNA")) +
      labs(x = "vOTU group",
           y = "Proportion of vOTUs\nwithin group",
           fill = "AMG carriage:") +
      theme(legend.position = "none",
            text = element_text(size = 8))
  
  right <- plot_grid(legend, plot, ncol = 1, rel_heights = c(1, 1))
  
  D <- plot_grid(left, right, nrow = 1, rel_widths = c(2, 1), labels = "D", label_size = 15)

  rm(list = c("nodes", "edges", "VC_nodes"))
  
```

```{r Supplementary Figure S1 - Taxonomic novelty of recovered soil vOTUs}

  plot <- plot_grid(A, B, C, D, ncol = 1, rel_heights = c(1, 1, 1, 1))

  ggsave(plot, width = 170, height = 340, units = "mm", dpi = 300,
         file = here("Figures/Figure S1.png"), device = "png")
  
```

# Supplementary Table S3 - Sources of hits to viral sequences in PhageClouds database

```{r Supplementary Table S3 - Sources of hits to viral sequences in PhageClouds database}

  PhageClouds_hits_sources <- 
    as_tibble(read.csv(here("Data/PhageClouds_hits_sources.csv")))

  write.csv(PhageClouds_hits_sources, file = here("Tables/Table S3.csv"), row.names = F)

```

# Supplementary Figure S2 - Phylogenetic assessment of jumbo phage vOTUs and jumbo-related vOTUs using DNA polymerase gene

```{r Supplementary Figure S2 - Phylogenetic assessment of jumbo phage vOTUs and jumbo-related vOTUs using DNA polymerase gene}

# import DNAP tree file
  tree <- 
    read.tree(file = here("Data/DNAP_tree"))
    
# root the tree on outgroup (BAM29052)
  tree <- ape::root(tree, outgroup = "BAM29052")
  
# import metadata on inphared reference genomes
  inphared_data <- 
    as_tibble(read.delim(here("Data/inphared_data_1Aug2022.tsv"))) %>%
    select("NAME" = Accession, "DESCRIPTION" = Description, "LENGTH" = `Genome.Length..bp.`, "HOST" = Host, "FAMILY" = Family, "GENUS" = Genus) %>% 
    mutate(FAMILY = case_when(FAMILY == "Myoviridae" ~ "Former Myoviridae",
                              FAMILY == "Podoviridae" ~ "Former Podoviridae",
                              FAMILY == "Siphoviridae" ~ "Former Siphoviridae",
                              TRUE ~ FAMILY))
# create gene contig index
  gene_contig_index <- 
    DNA_vOTU_gene_annotations %>% 
    select(NAME, LOCUS) %>% 
    left_join(select(DNA_vOTU_reads, c("NAME", "LENGTH")), by = "NAME")
  
# create annotation metadata data frame
  metadata <- 
    tibble(LOCUS = tree$tip.label) %>% 
    left_join(gene_contig_index, by = "LOCUS") %>% 
    mutate(NAME = case_when(is.na(NAME) ~ word(LOCUS, start = 1, end = 1, sep = "_"),
                            LOCUS == "BAM29052" ~ "Herpesvirus",
                            TRUE ~ NAME)) %>% 
    # correct names that have been updated
    mutate(NAME = case_when(NAME == "CAKLQF010000001" ~ "CAKLQF020000001",
                            NAME == "CAKLQF010000017" ~ "CAKLQF020000017",
                            NAME == "CAKLQF010000032" ~ "CAKLQF020000032",
                            NAME == "NC" ~ "NC_004820",
                            TRUE ~ NAME)) %>%
    left_join(inphared_data, by = "NAME") %>% 
    # re-label 50 unverifieed genomes missing in inphared metadata file
    mutate(LABEL = case_when(NAME == "MT835496" ~ "MT835496",
                             NAME == "MT835570" ~ "MT835570",
                             NAME == "MT836180" ~ "MT836180",
                             NAME == "MT835501" ~ "MT835501",
                             NAME == "MT836494" ~ "MT836494",
                             NAME == "MT836213" ~ "MT836213",
                             NAME == "MT835604" ~ "MT835604",
                             NAME == "MT835885" ~ "MT835885",
                             NAME == "MT835632" ~ "MT835632",
                             NAME == "MT836290" ~ "MT836290",
                             NAME == "MT836081" ~ "MT836081",
                             NAME == "MT835750" ~ "MT835750",
                             NAME == "MT835961" ~ "MT835961",
                             NAME == "MT836709" ~ "MT836709",
                             NAME == "MT835601" ~ "MT835601",
                             NAME == "MT835494" ~ "MT835494",
                             NAME == "MT835496" ~ "MT835496",
                             NAME == "MT836173" ~ "MT836173",
                             NAME == "JQ067083" ~ "Pseudomonas phage PaMx13",
                             NAME == "MT835953" ~ "MT835953",
                             NAME == "MT835981" ~ "MT835981",
                             NAME == "MT836050" ~ "MT836050",
                             NAME == "MT835481" ~ "MT835481",
                             NAME == "MT835556" ~ "MT835556",
                             NAME == "MT835764" ~ "MT835764",
                             NAME == "MT835747" ~ "MT835747",
                             NAME == "MW495066" ~ "Microcystis phage MaAM05",
                             NAME == "MT835640" ~ "MT835640",
                             NAME == "MT838268" ~ "MT838268",
                             NAME == "MT835490" ~ "MT835490",
                             NAME == "MT835512" ~ "MT835512",
                             NAME == "MT835753" ~ "MT835753",
                             NAME == "MT836467" ~ "MT836467",
                             NAME == "MT838720" ~ "MT838720",
                             NAME == "MT836618" ~ "MT836618",
                             NAME == "MT835881" ~ "MT835881",
                             NAME == "MT837764" ~ "MT837764",
                             NAME == "MT835891" ~ "MT835891",
                             NAME == "MT836087" ~ "MT836087",
                             NAME == "MT835911" ~ "MT835911",
                             NAME == "MT835482" ~ "MT835482",
                             NAME == "MN905546" ~ "Pseudomonas phage PN05",
                             NAME == "KT319599" ~ "Listeria phage WIL-2",
                             NAME == "MT836181" ~ "MT836181",
                             NAME == "MT835536" ~ "MT835536",
                             NAME == "MT835890" ~ "MT835890",
                             NAME == "MT836025" ~ "MT836025",
                             NAME == "MT836709" ~ "MT836709",
                             TRUE ~ as.character(NA))) %>%
    mutate(LABEL = case_when(is.na(LABEL) & grepl("DNA", NAME) ~ NAME,
                             is.na(LABEL) ~ DESCRIPTION,
                             NAME == "Herpesvirus" ~ "Human alphaherpesvirus 1",
                             TRUE ~ LABEL)) %>% 
    mutate(LENGTH = case_when(is.na(LENGTH.x) ~ LENGTH.y,
                              is.na(LENGTH.y) ~ LENGTH.x)) %>% 
    mutate(HOST = case_when(grepl("DNA_vOTU", NAME) ~ "vOTU",
                            is.na(HOST) ~ "Unknown",
                            TRUE ~ HOST)) %>% 
    mutate(FAMILY = case_when(grepl("DNA_vOTU", NAME) ~ "vOTU",
                              is.na(FAMILY) ~ "Unknown",
                              FAMILY == "Unclassified" ~ "Unknown",
                              TRUE ~ FAMILY)) %>% 
    mutate(GENUS = case_when(grepl("DNA_vOTU", NAME) ~ "vOTU",
                             is.na(GENUS) ~ "Unknown",
                             GENUS == "Unclassified" ~ "Unknown",
                             TRUE ~ GENUS)) %>% 
    select(LOCUS, NAME, LABEL, LENGTH, HOST, FAMILY, GENUS)
  
# save circular phylogenetic tree
  plot <-
    ggtree(tree, colour = "grey50", size = 0.2) %<+%
      metadata +
      layout_circular()

# add annotation layers to phylogenetic tree
  plot2 <-
    plot +
    geom_nodepoint(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) >= 70 & as.numeric(label) < 80), 
                   size = 0.2, colour = "grey50", alpha = 0.3) +
    geom_nodepoint(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) >= 80 & as.numeric(label) < 90), 
                   size = 0.4, colour = "grey50", alpha = 0.3) +
    geom_nodepoint(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) >= 90), 
                   size = 0.6, colour = "grey50", alpha = 0.3) +
    geom_tiplab2(aes(label = LABEL, colour = grepl("DNA_vOTU", LABEL)), align = T, linetype = NA, size = 0.3, offset = 0.2) +
    geom_tiplab2(aes(label = NA, colour = grepl("DNA_vOTU", LABEL), alpha = grepl("DNA_vOTU", LABEL)), align = T, linetype = "dotted", linesize = 0.05, offset = 0.2) +
    geom_cladelabel(node = 1329, label = "A", fontsize = 5, offset = 4.4, barsize = 0, color = c("white", "black")) +
    geom_cladelabel(node = 1356, label = "B", fontsize = 5, offset = 4.6, barsize = 0, color = c("white", "black")) +
    geom_cladelabel(node = 370, label = "C", fontsize = 5, offset = 4.1, barsize = 0, color = c("white", "black")) +
    geom_cladelabel(node = 1847, label = "D", fontsize = 5, offset = 3.6, barsize = 0, color = c("white", "black")) +
    geom_cladelabel(node = 1281, label = "E", fontsize = 5, offset = 4.1, barsize = 0, color = c("white", "black")) +
    geom_cladelabel(node = 1872, label = "F", fontsize = 5, offset = 2.9, barsize = 0, color = c("white", "black")) +
    geom_treescale(width = 1, x = 3, fontsize = 3, offset = 4) +
    scale_colour_manual(values = c("grey50", "goldenrod"), 
                        guide = "none") +
    scale_alpha_manual(values = c(0.3, 1.0),
                       guide = "none")
  
# create labels for heatmap
  heatmap_labels <- metadata %>% unique() %>% select(LOCUS, FAMILY) %>% column_to_rownames("LOCUS")
# create colour scheme for heatmap
  heatmap_colours <- c("white", '#1f78b4', "dark olive green", "pink", '#e31a1c', "lime green", '#ff7f00', "cyan", '#6a3d9a', '#FFFF99', "grey90")
  names(heatmap_colours) <- c("vOTU", "Demerecviridae", "Former Myoviridae", "Herelleviridae", "Molycolviridae", "Peduoviridae", "Former Podoviridae", "Schitoviridae", "Former Siphoviridae", "Zierdtviridae", "Unknown")
  
# add heatmap layer to phylogenetic tree
  plot3 <- 
    gheatmap(plot2, heatmap_labels, offset = 1.4, color = NULL, colnames_position = "top", colnames_angle = 90, colnames_offset_y = 1, hjust = 0, font.size = 2, width = 0.05, colnames = F) +
      scale_fill_manual(values = heatmap_colours, 
                        breaks = c("vOTU", "Demerecviridae", "Former Myoviridae", "Herelleviridae",
                                   "Molycolviridae", "Peduoviridae", "Former Podoviridae", "Schitoviridae",
                                   "Former Siphoviridae", "Zierdtviridae", "Unknown"),
                        labels = c("vOTU", "*Demerecviridae*", "Former *Myoviridae*", "*Herelleviridae*",
                                   "*Molycolviridae*", "*Peduoviridae*", "Former *Podoviridae*", "*Schitoviridae*",
                                   "Former *Siphoviridae*", "*Zierdtviridae*", "Other/Unknown"),
                        name = "Phage Family:") +
      guides(fill = guide_legend(override.aes = list(colour = c("black", rep("white", 10))))) +
      theme(legend.position = "bottom",
            text = element_text(size = 10),
            legend.text = element_markdown())
  
# save as pdf 6 x 6 inches
  ggsave(plot3, width = 170, height = 225, units = "mm", dpi = 300,
         file = here("Figures/Figure S2.pdf"), device = cairo_pdf)
    
```

# Supplementary Figure S3 - Phylogenetic groups A, B, and F from assessment of jumbo phage vOTUs and jumbo-related vOTUs using DNA polymerase gene

```{r Supplementary Figure S3A - Phylogenetic groups A and B from assessment of jumbo phage vOTUs and jumbo-related vOTUs using DNA polymerase gene}

# update annotation metadata data frame
  metadata2 <-
    metadata %>% 
    mutate(LENGTH = case_when(NAME == "MT835496" ~ as.integer(159206),
                              NAME == "MT835570" ~ as.integer(32488),
                              NAME == "MT836180" ~ as.integer(15601),
                              NAME == "MT835501" ~ as.integer(54463),
                              NAME == "MT836494" ~ as.integer(10059),
                              NAME == "MT836213" ~ as.integer(14083),
                              NAME == "MT835604" ~ as.integer(35036),
                              NAME == "MT835885" ~ as.integer(15914),
                              NAME == "MT835632" ~ as.integer(23709),
                              NAME == "MT836290" ~ as.integer(11336),
                              NAME == "MT836081" ~ as.integer(13142),
                              NAME == "MT835750" ~ as.integer(19297),
                              NAME == "MT835961" ~ as.integer(31546),
                              NAME == "MT836709" ~ as.integer(12256),
                              NAME == "MT835494" ~ as.integer(12256),
                              NAME == "MT835601" ~ as.integer(25781),
                              NAME == "MT835496" ~ as.integer(159206),
                              NAME == "MT836173" ~ as.integer(12185),
                              NAME == "JQ067083" ~ as.integer(66450),
                              NAME == "MT835953" ~ as.integer(18731),
                              NAME == "MT835981" ~ as.integer(14289),
                              NAME == "MT836050" ~ as.integer(13488),
                              NAME == "MT835481" ~ as.integer(74734),
                              NAME == "MT835556" ~ as.integer(38471),
                              NAME == "MT835764" ~ as.integer(18777),
                              NAME == "MT835747" ~ as.integer(19333),
                              NAME == "MW495066" ~ as.integer(273876),
                              NAME == "MT835640" ~ as.integer(23302),
                              NAME == "MT838268" ~ as.integer(10553),
                              NAME == "MT835490" ~ as.integer(60344),
                              NAME == "MT835512" ~ as.integer(44479),
                              NAME == "MT835753" ~ as.integer(19178),
                              NAME == "MT836467" ~ as.integer(10206),
                              NAME == "MT838720" ~ as.integer(5434),
                              NAME == "MT836618" ~ as.integer(18369),
                              NAME == "MT835881" ~ as.integer(22433),
                              NAME == "MT837764" ~ as.integer(11113),
                              NAME == "MT835891" ~ as.integer(15829),
                              NAME == "MT836087" ~ as.integer(13099),
                              NAME == "MT835911" ~ as.integer(18569),
                              NAME == "MT835482" ~ as.integer(72712),
                              NAME == "MN905546" ~ as.integer(214121),
                              NAME == "KT319599" ~ as.integer(27697),
                              NAME == "MT836181" ~ as.integer(16395),
                              NAME == "MT835536" ~ as.integer(40191),
                              NAME == "MT835890" ~ as.integer(15833),
                              NAME == "MT836025" ~ as.integer(13786),
                              NAME == "MT836709" ~ as.integer(12256),
                              TRUE ~ LENGTH)) %>% 
    mutate(LABEL = case_when(is.na(LABEL) & is.na(LENGTH) ~ as.character(NA),
                             TRUE ~ paste(LABEL, " (", format(LENGTH, big.mark = " ", trim = T), ")", sep = "")))

# save rectangular phylogenetic tree
  plot <-
    ggtree(tree, colour = "grey50", size = 0.7, layout = "rectangular") %<+%
    metadata2 +
    geom_text(aes(label = node), size = 0)

# visualise clade containing Groups A and B
  plot_AB <-
    viewClade(plot, node = 1326, xmax_adjust = -1.25) +
    geom_tiplab(aes(label = LABEL, colour = grepl("DNA_vOTU", LABEL)), align = T, linetype = NA, size = 1.8, offset = -1.8) +
    geom_tiplab(aes(label = NA, colour = grepl("DNA_vOTU", LABEL), alpha = grepl("DNA_vOTU", LABEL)), align = T, linetype = "dotted", linesize = 0.3, offset = -1.8) +
    geom_tippoint(aes(fill = FAMILY), shape = 21, size = 1.8, alpha = 0.6) +
    geom_cladelabel(node = 1329, label = "A", fontsize = 3.5, offset = 0.75, barsize = 0, hjust = -0.2) +
    geom_cladelabel(node = 1356, label = "B", fontsize = 3.5, offset = 0.90, barsize = 0) +
    geom_nodepoint(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) >= 70 & as.numeric(label) < 80),
                   size = 0.5, colour = "grey50", alpha = 0.4) +
    geom_nodepoint(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) >= 80 & as.numeric(label) < 90),
                   size = 1, colour = "grey50", alpha = 0.4) +
    geom_nodepoint(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) >= 90),
                   size = 1.5, colour = "grey50", alpha = 0.4) +
    geom_treescale(width = 0.5, fontsize = 3.5, y = 220, x = 1.85, offset = 2) +
    scale_fill_manual(values = heatmap_colours, 
                      breaks = c("vOTU", "Other/Unknown", "Former Myoviridae", "Former Siphoviridae", "Former Podoviridae", "Zierdtviridae", "Peduoviridae", "Molycolviridae", "Herelleviridae", "Demerecviridae", "Schitoviridae"),
                      name = "Phage Family:") +
    scale_colour_manual(values = c("grey50", "goldenrod"),
                        guide = "none") +
    scale_alpha_manual(values = c(0.3, 1.0),
                       guide = "none") +
    theme(legend.position = "none")

```

```{r Supplementary Figure S3B - Phylogenetic group F from assessment of jumbo phage vOTUs and jumbo-related vOTUs using DNA polymerase gene}

# update annotation metadata data frame
  metadata2 <-
    metadata %>% 
    mutate(LENGTH = case_when(NAME == "MT835496" ~ as.integer(159206),
                              NAME == "MT835570" ~ as.integer(32488),
                              NAME == "MT836180" ~ as.integer(15601),
                              NAME == "MT835501" ~ as.integer(54463),
                              NAME == "MT836494" ~ as.integer(10059),
                              NAME == "MT836213" ~ as.integer(14083),
                              NAME == "MT835604" ~ as.integer(35036),
                              NAME == "MT835885" ~ as.integer(15914),
                              NAME == "MT835632" ~ as.integer(23709),
                              NAME == "MT836290" ~ as.integer(11336),
                              NAME == "MT836081" ~ as.integer(13142),
                              NAME == "MT835750" ~ as.integer(19297),
                              NAME == "MT835961" ~ as.integer(31546),
                              NAME == "MT836709" ~ as.integer(12256),
                              NAME == "MT835494" ~ as.integer(12256),
                              NAME == "MT835601" ~ as.integer(25781),
                              NAME == "MT835496" ~ as.integer(159206),
                              NAME == "MT836173" ~ as.integer(12185),
                              NAME == "JQ067083" ~ as.integer(66450),
                              NAME == "MT835953" ~ as.integer(18731),
                              NAME == "MT835981" ~ as.integer(14289),
                              NAME == "MT836050" ~ as.integer(13488),
                              NAME == "MT835481" ~ as.integer(74734),
                              NAME == "MT835556" ~ as.integer(38471),
                              NAME == "MT835764" ~ as.integer(18777),
                              NAME == "MT835747" ~ as.integer(19333),
                              NAME == "MW495066" ~ as.integer(273876),
                              NAME == "MT835640" ~ as.integer(23302),
                              NAME == "MT838268" ~ as.integer(10553),
                              NAME == "MT835490" ~ as.integer(60344),
                              NAME == "MT835512" ~ as.integer(44479),
                              NAME == "MT835753" ~ as.integer(19178),
                              NAME == "MT836467" ~ as.integer(10206),
                              NAME == "MT838720" ~ as.integer(5434),
                              NAME == "MT836618" ~ as.integer(18369),
                              NAME == "MT835881" ~ as.integer(22433),
                              NAME == "MT837764" ~ as.integer(11113),
                              NAME == "MT835891" ~ as.integer(15829),
                              NAME == "MT836087" ~ as.integer(13099),
                              NAME == "MT835911" ~ as.integer(18569),
                              NAME == "MT835482" ~ as.integer(72712),
                              NAME == "MN905546" ~ as.integer(214121),
                              NAME == "KT319599" ~ as.integer(27697),
                              NAME == "MT836181" ~ as.integer(16395),
                              NAME == "MT835536" ~ as.integer(40191),
                              NAME == "MT835890" ~ as.integer(15833),
                              NAME == "MT836025" ~ as.integer(13786),
                              NAME == "MT836709" ~ as.integer(12256),
                              TRUE ~ LENGTH)) %>% 
    mutate(LABEL = case_when(is.na(LABEL) & is.na(LENGTH) ~ as.character(NA),
                             TRUE ~ paste(LABEL, " (", format(LENGTH, big.mark = " ", trim = T), ")", sep = "")))

# save rectangular phylogenetic tree
  plot <-
    ggtree(tree, colour = "grey50", size = 0.7, layout = "rectangular") %<+%
    metadata2 +
    geom_text(aes(label = node), size = 0)

# visualise clade containing Group F
  plot_F <-
    viewClade(plot, node = 1868, xmax_adjust = 0.6) +
    geom_tiplab(aes(label = LABEL, colour = grepl("DNA_vOTU", LABEL)), align = T, linetype = NA, size = 1.8, offset = -0.2) +
    geom_tiplab(aes(label = NA, colour = grepl("DNA_vOTU", LABEL), alpha = grepl("DNA_vOTU", LABEL)), align = T, linetype = "dotted", linesize = 0.3, offset = -0.2) +
    geom_tippoint(aes(fill = FAMILY), shape = 21, size = 1.8, alpha = 0.6) +
    geom_cladelabel(node = 1872, label = "F", fontsize = 3.5, offset = 0.8, barsize = 0) +
    geom_nodepoint(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) >= 70 & as.numeric(label) < 80),
                   size = 0.5, colour = "grey50", alpha = 0.4) +
    geom_nodepoint(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) >= 80 & as.numeric(label) < 90),
                   size = 1, colour = "grey50", alpha = 0.4) +
    geom_nodepoint(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) >= 90),
                   size = 1.5, colour = "grey50", alpha = 0.4) +
    geom_treescale(width = 0.5, fontsize = 3.5, y = 742, x = 3.2, offset = 0.6) +
    scale_fill_manual(values = heatmap_colours, 
                      breaks = c("vOTU", "Other/Unknown", "Former Myoviridae", "Former Siphoviridae", "Former Podoviridae", "Zierdtviridae", "Peduoviridae", "Molycolviridae", "Herelleviridae", "Demerecviridae", "Schitoviridae"),
                      name = "Phage Family:") +
    scale_colour_manual(values = c("grey50", "goldenrod"),
                        guide = "none") +
    scale_alpha_manual(values = c(0.3, 1.0),
                       guide = "none") +
    theme(legend.position = "none")

```

```{r Supplementary Figure S3 - Phylogenetic groups A, B, and F from assessment of jumbo phage vOTUs and jumbo-related vOTUs using DNA polymerase gene}

  plot <- plot_grid(plot_AB, plot_F, labels = c("A", "B"), ncol = 1, rel_heights = c(0.75, 0.25), label_size = 15)

  ggsave(plot, width = 170, height = 225, units = "mm", dpi = 300,
         file = here("Figures/Figure S3.pdf"), device = cairo_pdf)

  # ggsave(plot_AB, width = 170, height = 225, units = "mm", dpi = 300,
  #        file = here("Figures/Figure S3A.pdf"), device = cairo_pdf)

  # ggsave(plot_F, width = 170, height = 225, units = "mm", dpi = 300,
  #        file = here("Figures/Figure S3B.pdf"), device = cairo_pdf)

```

# Figure 1 - Viral distribution throughout soil depth profiles

```{r Figure 1 - Overview of viral community structure throughout soil depth profiles}

  df <-
    DNA_vOTU_reads %>% 
    select(NAME, VC_STATUS, CLUSTER, contains("Sage")) %>% 
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "READS") %>% 
    filter(READS > 0) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                             grepl("100", SAMPLE) ~ 100,
                             grepl("80", SAMPLE) ~ 80,
                             grepl("60", SAMPLE) ~ 60,
                             grepl("40", SAMPLE) ~ 40,
                             grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))

  input <-
    df %>% 
    filter(SITE == "Hilly grassland") %>% 
    select(NAME, DEPTH) %>% 
    unique() %>% 
    mutate(PRESENCE = 1) %>% 
    pivot_wider(names_from = "DEPTH", values_from = "PRESENCE", values_fill = 0)

  top <-
    upset(data = input, 
          intersect = c("115", "100", "80", "60", "40", "20"),
          name = "Sample intersection",
          set_sizes = F,
          base_annotations = list(
            "vOTUs in intersection" = intersection_ratio(#mapping = aes(fill = "bars_colour"), 
                                                         text_mapping = aes(label = paste0(!!upset_text_percentage(digits = 1), "\n", "(", !!get_size_mode("exclusive_intersection"), ")")),
                                                                        text = list(size = 2.5)) +
              # scale_fill_manual(values = c("bars_colour" = "black"), guide = "none") +
              scale_y_continuous(labels = scales::percent_format(),
                                 limits = c(0, 1)) + 
              facet_grid(rows = vars("Hilly grassland")) +
              ggtitle("")),
          matrix = (intersection_matrix(geom = geom_point(size = 1.5)) + 
                      ylab("Soil depth (cm)")),
          themes = upset_modify_themes(
            list(
              "vOTUs in intersection" = theme_bw() +
                theme(axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(),
                      text = element_text(size = 10)),
              "intersections_matrix" = theme_bw() + 
                theme(#axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(), 
                      axis.line.x = element_blank(),
                      text = element_text(size = 10)),
              "overall_sizes" = theme_bw() +
                theme(axis.title.y = element_blank(),
                      axis.text.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      axis.line.y = element_blank(),
                      text = element_text(size = 10))
              )
            ),
          stripes = c("grey95", "white"),
          width_ratio = 0.3,
          height_ratio = 0.4,
          sort_intersections_by = c("degree", "cardinality"),
          sort_intersections = c("descending"),
          group_by = "degree",
          sort_sets = FALSE,
          min_size = 60)
  
  input <-
    df %>% 
    filter(SITE == "Garry Oak") %>% 
    select(NAME, DEPTH) %>% 
    unique() %>% 
    mutate(PRESENCE = 1) %>% 
    pivot_wider(names_from = "DEPTH", values_from = "PRESENCE", values_fill = 0)
  
  bottom <-
    upset(data = input, 
          intersect = c("115", "100", "80", "60", "40", "20"),
          name = "Sample intersection",
          set_sizes = F,
          base_annotations = list(
            "vOTUs in intersection" = intersection_ratio(#mapping = aes(fill = "bars_colour"), 
                                                         text_mapping = aes(label = paste0(!!upset_text_percentage(digits = 1), "\n", "(", !!get_size_mode("exclusive_intersection"), ")")),
                                                                        text = list(size = 2.5)) +
              #scale_fill_manual(values = c("bars_colour" = "black"), guide = "none") +
              scale_y_continuous(labels = scales::percent_format(),
                                 limits = c(0, 1)) +
              facet_grid(rows = vars("Garry Oak"))),
          matrix = (intersection_matrix(geom = geom_point(size = 1.5)) + 
                      ylab("Soil depth (cm)")),
          themes = upset_modify_themes(
            list(
              "vOTUs in intersection" = theme_bw() +
                theme(axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(),
                      text = element_text(size = 10)),
              "intersections_matrix" = theme_bw() + 
                theme(#axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(), 
                      axis.line.x = element_blank(),
                      text = element_text(size = 10)),
              "overall_sizes" = theme_bw() +
                theme(axis.title.y = element_blank(),
                      axis.text.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      axis.line.y = element_blank(),
                      text = element_text(size = 10))
              )
            ),
          stripes = c("grey95", "white"),
          width_ratio = 0.3,
          height_ratio = 0.4,
          sort_intersections_by = c("degree", "cardinality"),
          sort_intersections = c("descending"),
          group_by = "degree",
          sort_sets = FALSE,
          min_size = 60)
  
  A <- plot_grid(top, bottom, ncol = 1, rel_heights = c(0.52, 0.48), labels = "A", label_size = 15)
  
  input1 <- 
    df %>% 
    group_by(NAME, DEPTH, SITE) %>% 
    summarise(PRESENCE = case_when(n() == 2 ~ "BOTH REPS",
                                   n() == 1 ~ "ONE REP"), .groups = "drop") %>% 
    filter(PRESENCE == "BOTH REPS") %>% 
    group_by(SITE, DEPTH) %>% 
    summarise(NUM_vOTUs = n(),
              PROP_vOTUs = (n()/9787), .groups = "drop")
  
  input2 <- 
    df %>% 
    group_by(NAME, DEPTH) %>% 
    summarise(PRESENCE = case_when(n() == 3 ~ "Three replicates",
                                   n() == 4 ~ "Four replicates"), .groups = "drop") %>% 
    filter(!is.na(PRESENCE)) %>% 
    group_by(PRESENCE, DEPTH) %>% 
    summarise(NUM_vOTUs = n(),
              PROP_vOTUs = n()/9789, .groups = "drop") %>% 
    mutate(PRESENCE = factor(PRESENCE, levels = c("Three replicates", "Four replicates"))) %>% 
    filter(PRESENCE == "Four replicates") %>% 
    mutate(SITE = "Both sites") %>% 
    select(SITE, DEPTH, NUM_vOTUs, PROP_vOTUs)
  
  input <- 
    bind_rows(input1, input2) %>% 
    mutate(SITE = factor(SITE, levels = c("Hilly grassland", "Garry Oak", "Both sites")))

  plot <- 
    ggplot(input, aes(x = DEPTH, y = PROP_vOTUs)) +
      geom_bar(stat = "identity") +
      # geom_text(aes(label = paste(format(100*PROP_vOTUs, digits = 3), "%\n(", NUM_vOTUs, ")", sep = "")), 
      #           vjust = 2, colour = "white", size = 3) +
      facet_grid(rows = vars(SITE)) +
      scale_x_continuous(limits = c(10, 125),
                         breaks = c(20, 40, 60, 80, 100, 115)) +
      scale_y_continuous(labels = scales::percent_format(),
                                   limits = c(0, 1)) +
      labs(x = "Soil depth (cm)",
           y = "vOTUs shared between replicates",
           title = "")

  B <- plot_grid(plot, ncol = 1, labels = "B", label_size = 15)
  
  input <- 
    df %>% 
    group_by(NAME, DEPTH) %>% 
    mutate(PRESENCE = case_when(n() == 4 ~ "Four replicates",
                                TRUE ~ "NA")) %>% 
    group_by(PRESENCE, DEPTH) %>% 
    summarise(VIRAL_READS = sum(READS), .groups = "drop") %>% 
    group_by(DEPTH) %>% 
    mutate(PROP_READS = VIRAL_READS/sum(VIRAL_READS)) %>% 
    filter(PRESENCE == "Four replicates")

  plot <- 
    ggplot(input, aes(x = DEPTH, y = PROP_READS)) +
      geom_bar(stat = "identity") +
      geom_text(aes(label = paste(format(100*PROP_READS, digits = 3), "%", sep = "")),
                vjust = 2, colour = "white", size = 3) +
      facet_grid(rows = vars("Both sites")) +
      scale_x_continuous(limits = c(10, 125),
                         breaks = c(20, 40, 60, 80, 100, 115)) +
      scale_y_continuous(labels = scales::percent_format(),
                                   limits = c(0, 1)) +
      labs(x = "Soil depth (cm)",
           y = "Viral reads shared\nbetween replicates",
           title = "")
  
  C <- plot_grid(plot, ncol = 1, labels = "C", label_size = 15)
  
  top <- plot_grid(A, B, nrow = 1, rel_widths = c(0.5, 0.5))
  
  plot <- plot_grid(top, C, ncol = 1, rel_heights = c(0.7, 0.3))
  
  ggsave(plot, width = 170, height = 225, units = "mm", dpi = 300,
         file = here("Figures/Figure 1.eps"), device = cairo_ps,  fallback_resolution = 300)
  
```

# Figure S4 - Overview of microbial community structure throughout soil depth profiles

```{r Figure S4 - Overview of microbial community structure throughout soil depth profiles}

  df <-
    OTU_coverage %>% 
    select(NAME, contains("Sage")) %>% 
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "REL_ABUND") %>% 
    filter(REL_ABUND > 0) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                             grepl("100", SAMPLE) ~ 100,
                             grepl("80", SAMPLE) ~ 80,
                             grepl("60", SAMPLE) ~ 60,
                             grepl("40", SAMPLE) ~ 40,
                             grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))

  input <-
    df %>% 
    filter(SITE == "Hilly grassland") %>% 
    select(NAME, DEPTH) %>% 
    unique() %>% 
    mutate(PRESENCE = 1) %>% 
    pivot_wider(names_from = "DEPTH", values_from = "PRESENCE", values_fill = 0)

  top <-
    upset(data = input, 
          intersect = c("115", "100", "80", "60", "40", "20"),
          name = "",
          set_sizes = F,
          base_annotations = list(
            "OTUs in intersection" = intersection_ratio(mapping = aes(fill = "bars_colour"), text_mapping = aes(label = paste0(!!upset_text_percentage(digits = 1), "\n", "(", !!get_size_mode("exclusive_intersection"), ")")),
                                                                        text = list(size = 2.5)) +
              scale_fill_manual(values = c("bars_colour" = "black"), guide = "none") +
              scale_y_continuous(labels = scales::percent_format(),
                                 limits = c(0, 0.25),
                                 breaks = seq(0, 0.25, 0.05)) + 
              facet_grid(rows = vars("Hilly grassland")) +
              ggtitle("")),
          matrix = (intersection_matrix(geom = geom_point(size = 1.5)) + 
                      ylab("Soil depth (cm)")),
          themes = upset_modify_themes(
            list(
              "OTUs in intersection" = theme_bw() +
                theme(axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(),
                      text = element_text(size = 10)),
              "intersections_matrix" = theme_bw() + 
                theme(axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(), 
                      axis.line.x = element_blank(),
                      text = element_text(size = 10)),
              "overall_sizes" = theme_bw() +
                theme(axis.title.y = element_blank(),
                      axis.text.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      axis.line.y = element_blank(),
                      text = element_text(size = 10))
              )
            ),
          stripes = c("grey95", "white"),
          width_ratio = 0.3,
          height_ratio = 0.4,
          sort_intersections_by = c("degree", "cardinality"),
          sort_intersections = c("descending"),
          group_by = "degree",
          sort_sets = FALSE,
          min_size = 74)
  
  input <-
    df %>% 
    filter(SITE == "Garry Oak") %>% 
    select(NAME, DEPTH) %>% 
    unique() %>% 
    mutate(PRESENCE = 1) %>% 
    pivot_wider(names_from = "DEPTH", values_from = "PRESENCE", values_fill = 0)
  
  bottom <-
    upset(data = input, 
          intersect = c("115", "100", "80", "60", "40", "20"),
          name = "",
          set_sizes = F,
          base_annotations = list(
            "OTUs in intersection" = intersection_ratio(mapping = aes(fill = "bars_colour"), text_mapping = aes(label = paste0(!!upset_text_percentage(digits = 1), "\n", "(", !!get_size_mode("exclusive_intersection"), ")")),
                                                                        text = list(size = 2.5)) +
              scale_fill_manual(values = c("bars_colour" = "black"), guide = "none") +
              scale_y_continuous(labels = scales::percent_format(),
                                 limits = c(0, 0.25),
                                 breaks = seq(0, 0.25, 0.05)) + 
              facet_grid(rows = vars("Garry Oak")) +
              ggtitle("")),
          matrix = (intersection_matrix(geom = geom_point(size = 1.5)) + 
                      ylab("Soil depth (cm)")),
          themes = upset_modify_themes(
            list(
              "OTUs in intersection" = theme_bw() +
                theme(axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(),
                      text = element_text(size = 10)),
              "intersections_matrix" = theme_bw() + 
                theme(axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(), 
                      axis.line.x = element_blank(),
                      text = element_text(size = 10)),
              "overall_sizes" = theme_bw() +
                theme(axis.title.y = element_blank(),
                      axis.text.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      axis.line.y = element_blank(),
                      text = element_text(size = 10))
              )
            ),
          stripes = c("grey95", "white"),
          width_ratio = 0.3,
          height_ratio = 0.4,
          sort_intersections_by = c("degree", "cardinality"),
          sort_intersections = c("descending"),
          group_by = "degree",
          sort_sets = FALSE,
          min_size = 25)
  
  A <- plot_grid(top, bottom, ncol = 1, rel_heights = c(0.52, 0.48), labels = "A", label_size = 15)
  
  input1 <- 
    df %>% 
    group_by(NAME, DEPTH, SITE) %>% 
    summarise(PRESENCE = case_when(n() == 2 ~ "BOTH REPS",
                                   n() == 1 ~ "ONE REP"), .groups = "drop") %>% 
    filter(PRESENCE == "BOTH REPS") %>% 
    group_by(SITE, DEPTH) %>% 
    summarise(NUM_OTUs = n(),
              PROP_OTUs = (n()/1447), .groups = "drop")
  
  input2 <- 
    df %>% 
    group_by(NAME, DEPTH) %>% 
    summarise(PRESENCE = case_when(n() == 3 ~ "Three replicates",
                                   n() == 4 ~ "Four replicates"), .groups = "drop") %>% 
    filter(!is.na(PRESENCE)) %>% 
    group_by(PRESENCE, DEPTH) %>% 
    summarise(NUM_OTUs = n(),
              PROP_OTUs = n()/1447, .groups = "drop") %>% 
    mutate(PRESENCE = factor(PRESENCE, levels = c("Three replicates", "Four replicates"))) %>% 
    filter(PRESENCE == "Four replicates") %>% 
    mutate(SITE = "Both sites") %>% 
    select(SITE, DEPTH, NUM_OTUs, PROP_OTUs)
  
  input <- 
    bind_rows(input1, input2) %>% 
    mutate(SITE = factor(SITE, levels = c("Hilly grassland", "Garry Oak", "Both sites")))

  plot <-
    ggplot(input, aes(x = DEPTH, y = PROP_OTUs)) +
      geom_bar(stat = "identity") +
      # geom_text(aes(label = paste(format(100*PROP_vOTUs, digits = 3), "%\n(", NUM_vOTUs, ")", sep = "")), 
      #           vjust = 2, colour = "white", size = 3) +
      facet_grid(rows = vars(SITE)) +
      scale_x_continuous(limits = c(10, 125),
                         breaks = c(20, 40, 60, 80, 100, 115)) +
      scale_y_continuous(labels = scales::percent_format(),
                                   limits = c(0, 0.4)) +
      labs(x = "Soil depth (cm)",
           y = "OTUs shared between replicates",
           title = "")

  B <- plot_grid(plot, ncol = 1, labels = "B", label_size = 15)
  
  input <- 
    df %>% 
    group_by(NAME, DEPTH) %>% 
    mutate(PRESENCE = case_when(n() == 4 ~ "Four replicates",
                                TRUE ~ "NA")) %>% 
    group_by(PRESENCE, DEPTH) %>% 
    summarise(MICROBIAL_ABUNDANCE = sum(REL_ABUND), .groups = "drop") %>% 
    group_by(DEPTH) %>% 
    mutate(PROP_ABUNDANCE = MICROBIAL_ABUNDANCE/sum(MICROBIAL_ABUNDANCE)) %>% 
    filter(PRESENCE == "Four replicates")

  plot <-
    ggplot(input, aes(x = DEPTH, y = PROP_ABUNDANCE)) +
      geom_bar(stat = "identity") +
      geom_text(aes(label = paste(format(100*PROP_ABUNDANCE, digits = 3), "%", sep = "")),
                vjust = 2, colour = "white", size = 3) +
      facet_grid(rows = vars("Both sites")) +
      scale_x_continuous(limits = c(10, 125),
                         breaks = c(20, 40, 60, 80, 100, 115)) +
      scale_y_continuous(labels = scales::percent_format(),
                         limits = c(0, 0.75),
                         breaks = seq(0, 0.75, 0.25)) +
      labs(x = "Soil depth (cm)",
           y = "Microbial reads shared\nbetween replicates",
           title = "")
  
  C <- plot_grid(plot, ncol = 1, labels = "C", label_size = 15)
  
  top <- plot_grid(A, B, nrow = 1, rel_widths = c(0.5, 0.5))
  
  plot <- plot_grid(top, C, ncol = 1, rel_heights = c(0.7, 0.3))
  
  ggsave(plot, width = 170, height = 225, units = "mm", dpi = 300,
         file = here("Figures/Figure S4.pdf"), device = "pdf")
  
```

# Figure 2 - Population-level assembly of soil viral communities throughout soil depth

```{r OLD Figure 2A - Alpha diversity of viral communities}

# calculate vOTU CPM
  CPM <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    mutate(across(contains("Sage"), ~round(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME")

# generate phyloSeq vOTU matrix
  phyloseq <- otu_table(CPM, taxa_are_rows = T)

# create table of observed vOTU abundance and alpha diversity measures 
  alpha_diversity <-
    estimate_richness(phyloseq) %>%
    as_tibble(., rownames = "SAMPLE") %>%
    mutate(Pielou = Shannon / log(Observed)) %>%
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))

# create data frame for plotting 
  df <- 
    alpha_diversity %>% 
    select(SAMPLE, "Richness" = Observed, "Evenness" = Pielou, "Diversity" = Shannon, DEPTH, SITE) %>% 
    pivot_longer(c(Richness, Evenness, Diversity), names_to = "METRIC", values_to = "VALUE") %>% 
    mutate(METRIC = factor(METRIC, levels = c("Richness", "Evenness", "Diversity")))
  
# compute correlation coefficients for alpha diversity across soil depth
  stat_correlation <-
    df %>%
    nest(data = c(-SITE, -METRIC)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$DEPTH, .x$VALUE, method = "pearson")),
           lm_test = map(data, ~lm(.x$VALUE ~ .x$DEPTH)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    filter(cor_tidied.p.value < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, METRIC, label) %>% 
    mutate(XPOS = c(30, 90, 90, 90),
           YPOS = c(7500, 8000, 0.83, 7.5))
    
# visualise alpha diversity of viral communities
  plot <-
    ggplot(df, aes(x = DEPTH, y = VALUE)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(data = filter(df, paste(SITE, METRIC) %in% paste(stat_correlation$SITE, stat_correlation$METRIC)),
                  colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 3, lineheight = 1) +
      facet_grid(rows = vars(METRIC), cols = vars(SITE),
                 scales = "free_y") +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                         labels = c("Hilly grassland", "Garry Oak"), 
                         values = c(15, 17),
                         guide = "none") + 
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)", 
           y = "",
           title = "")
  
  A <- plot_grid(plot, labels = "A", label_size = 15)
  
```

```{r Figure 2A - Alpha diversity of viral communities}

# calculate vOTU CPM
  CPM <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    mutate(across(contains("Sage"), ~round(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME")

# generate phyloSeq vOTU matrix
  phyloseq <- otu_table(CPM, taxa_are_rows = T)

# create table of observed vOTU abundance and alpha diversity measures 
  alpha_diversity <-
    estimate_richness(phyloseq) %>%
    as_tibble(., rownames = "SAMPLE") %>%
    mutate(Pielou = Shannon / log(Observed)) %>%
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))

# import total library reads for each sample
  total_reads <-
    read_excel(here("../library_sizes.xlsx"))

# compute total viral reads for each sample
  viral_reads <-
    DNA_vOTU_reads %>%
    select(NAME, contains("Sage")) %>% 
    pivot_longer(cols = where(is.numeric), names_to = "SAMPLE", values_to = "READS") %>% 
    group_by(SAMPLE) %>% 
    summarise(VIRAL_READS = sum(READS)) %>% 
    left_join(total_reads, by = "SAMPLE") %>% 
    mutate(PROP_VIRAL_READS = VIRAL_READS/TOTAL_READS) %>% 
    rename(`Viral reads` = PROP_VIRAL_READS)

# compute Berger-Parker index for each sample
  bp <- 
    CPM %>% 
    t() %>% 
    as_tibble(rownames = "SAMPLE") %>% 
    pivot_longer(cols = where(is.numeric), names_to = "NAME", values_to = "CPM") %>% 
    group_by(SAMPLE) %>% 
    mutate(REL_ABUND = CPM/sum(CPM)) %>% 
    summarise(REL_ABUND = max(REL_ABUND)) %>% 
    rename(`Berger-Parker` = REL_ABUND)

# create data frame for plotting 
  df <-
    alpha_diversity %>% 
    left_join(bp, by = "SAMPLE") %>% 
    left_join(viral_reads, by = "SAMPLE") %>% 
    select(SAMPLE, "Viral reads", "Evenness" = Pielou, "Diversity" = Simpson, "Dominance" = `Berger-Parker`, DEPTH, SITE) %>% 
    pivot_longer(c(`Viral reads`, Evenness, Diversity, Dominance), names_to = "METRIC", values_to = "VALUE") %>% 
    mutate(METRIC = factor(METRIC, levels = c("Viral reads", "Evenness", "Diversity", "Dominance")))
  
# compute correlation coefficients for alpha diversity across soil depth
  stat_correlation <-
    df %>%
    nest(data = c(-SITE, -METRIC)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$DEPTH, .x$VALUE, method = "pearson")),
           lm_test = map(data, ~lm(.x$VALUE ~ .x$DEPTH)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    filter(cor_tidied.p.value < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, METRIC, label) %>% 
    mutate(XPOS = c(25, 90, 90),
           YPOS = c(0.045, 0.83, 0.998))
    
# visualise alpha diversity of viral communities
  plot <-
    ggplot(df, aes(x = DEPTH, y = VALUE)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(data = filter(df, paste(SITE, METRIC) %in% paste(stat_correlation$SITE, stat_correlation$METRIC)),
                  colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 2.5, lineheight = 1) +
      facet_grid(rows = vars(METRIC), cols = vars(SITE),
                 scales = "free_y") +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                         labels = c("Hilly grassland", "Garry Oak"), 
                         values = c(15, 17),
                         guide = "none") + 
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)", 
           y = "",
           title = "")
    
  A <- plot_grid(plot, labels = "A", label_size = 15)
    
```

```{r Figure 2B - Beta diversity of viral communities}

# calculate vOTU CPM
  CPM <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    mutate(across(contains("Sage"), ~sqrt(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME") %>% 
    t()

# compute Bray-Curtis dissimilarity on CPM
  CPM_dist <- vegdist(CPM, method = "bray")

# create metadata table
  metadata <- 
    tibble(SAMPLE = rownames(as.matrix(CPM_dist))) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ "115 cm",
                             grepl("100", SAMPLE) ~ "100 cm",
                             grepl("80", SAMPLE) ~ "80 cm",
                             grepl("60", SAMPLE) ~ "60 cm",
                             grepl("40", SAMPLE) ~ "40 cm",
                             grepl("20", SAMPLE) ~ "20 cm")) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))

# calculate NMDS ordination
  MDS <- metaMDS(CPM_dist,
                 distance = "bray",
                 autotransform = F,
                 k = 2,
                 trace = F)

# create data frame of sample coordinates
  df <-
    as_tibble(MDS$points, rownames = "SAMPLE") %>%
    left_join(metadata, by = "SAMPLE")
  
# create column metadata table
  metadata <- 
    tibble(SAMPLE = rownames(as.matrix(CPM_dist))) %>%
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                              grepl("100", SAMPLE) ~ 100,
                              grepl("80", SAMPLE) ~ 80,
                              grepl("60", SAMPLE) ~ 60,
                              grepl("40", SAMPLE) ~ 40,
                              grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(DEPTH = as.vector(scale(DEPTH)))
  
# run PERMANOVA on SITE and DEPTH factors
  permanova <- 
    adonis2(CPM_dist~SITE + DEPTH, 
            data = metadata, 
            permutations = 999, 
            method = "bray", 
            by = "margin") %>%
    tidy() 
  
# plot NMDS coordinates
  plot <-
    ggplot(df, aes(MDS1,MDS2)) +
      geom_point(aes(fill = DEPTH, shape = SITE), 
                 size = 5, alpha = 1, stroke = 0) +
      annotate("text", x = -0.7, y = 0.4, size = 3.5, hjust = 0, lineheight = 1,
               label = paste("Stress =", format(round(MDS$stress, 2), nsmall = 2))) +
      annotate("text", x = -0.7, y = -0.3, size = 3.5, hjust = 0, lineheight = 1,
               label = paste("Site = ", signif(100*permanova$R2[1], 3), "%",
                             "\np = ", signif(permanova$p.value[1], 3), sep = "")) +
      annotate("text", x = -0.4, y = -0.3, size = 3.5, hjust = 0, lineheight = 1,
               label = paste("Soil depth = ", signif(100*permanova$R2[2], 3), "%",
                             "\np = ", signif(permanova$p.value[2], 3), sep = "")) +
      # coord_fixed() +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                         labels = c("Hilly grassland", "Garry Oak"),
                         values = c(22, 24)) +
      scale_fill_manual(breaks = c("20 cm", "40 cm", "60 cm", "80 cm", "100 cm", "115 cm"),
                        labels = c("20 cm", "40 cm", "60 cm", "80 cm", "100 cm", "115 cm"),
                        values = browns_palette) +
      scale_y_continuous(limits = c(-0.35, 0.45)) +
      guides(fill = guide_legend(override.aes = list(shape = 21)),
             shape = guide_legend(override.aes = list(fill = "white", stroke = 1, size = 3))) +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(),
            text = element_text(size = 10)) +
      labs(x = "Coordinate 1",
           y = "Coordinate 2",
           fill = "Soil depth:",
           shape = "Site:",
           title = "")
  
  B <- plot_grid(plot, labels = "B", label_size = 15)

```

```{r Figure 2C - Distance-decay relationship of viral community structure}

# calculate vOTU CPM
  CPM <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    mutate(across(contains("Sage"), ~sqrt(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME") %>% 
    t()

# compute Bray-Curtis dissimilarity on CPM
  CPM_dist <- vegdist(CPM, method = "bray")

# create data frame of pairwise distances
  df <- 
    CPM_dist %>% 
    dist2list(.) %>% 
    as_tibble() %>% 
    mutate(across(c(col, row), as.character)) %>% 
    filter(col < row) %>% 
    filter((grepl("Sage1", col) & grepl("Sage1", row)) |
           (grepl("Sage2", col) & grepl("Sage2", row))) %>%
    mutate(SITE = case_when(grepl("Sage1", col) & grepl("Sage1", row) ~ "Hilly grassland",
                            grepl("Sage2", col) & grepl("Sage2", row) ~ "Garry Oak")) %>% 
    mutate(DEPTH1 = case_when(grepl("113", col) | grepl("115", col) ~ 115,
                              grepl("100", col) ~ 100,
                              grepl("80", col) ~ 80,
                              grepl("60", col) ~ 60,
                              grepl("40", col) ~ 40,
                              grepl("20", col) ~ 20)) %>% 
    mutate(DEPTH2 = case_when(grepl("113", row) | grepl("115", row) ~ 115,
                              grepl("100", row) ~ 100,
                              grepl("80", row) ~ 80,
                              grepl("60", row) ~ 60,
                              grepl("40", row) ~ 40,
                              grepl("20", row) ~ 20)) %>% 
    mutate(DEPTH_DIFFERENCE = abs(DEPTH1 - DEPTH2)) %>% 
    rename(DISTANCE = value)

# compute correlation coefficients for pairwise distances and difference in soil depth
  stat_correlation <- 
    df %>% 
    nest(data = -SITE) %>%
    mutate(cor_test = map(data, ~cor.test(.x$DEPTH_DIFFERENCE, .x$DISTANCE, method = "pearson")),
           lm_test = map(data, ~lm(.x$DISTANCE ~ .x$DEPTH_DIFFERENCE)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, label) %>% 
    mutate(XPOS = c(70, 70),
           YPOS = c(0.17, 0.14))
  
# plot correlation plot
  plot <- 
    ggplot(df, aes(x = DEPTH_DIFFERENCE, y = DISTANCE)) +
      geom_point(aes(shape = SITE),
                 color = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                size = 3, hjust = 0, lineheight = 1) +
      facet_grid(cols = vars(SITE)) +
      scale_x_continuous(limits = c(0, 100),
                         breaks = seq(0, 100, 20)) +
      scale_y_continuous(limits = c(0, 0.6),
                         breaks = seq(0, 0.6, 0.1)) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                         labels = c("Hilly grassland", "Garry Oak"), 
                         values = c(15, 17),
                         guide = "none") + 
      theme(text = element_text(size = 10)) +
      labs(x = "Difference in soil depth (cm)",
           y = "Pairwise Bray-Curtis dissimilarity",
           title = "")
  
  C <- plot_grid(plot, labels = "C", label_size = 15)

```

```{r Figure 2 - Population-level assembly of soil viral communities throughout soil depth}

  plot <- plot_grid(A, B, C, ncol = 1, rel_heights = c(0.4, 0.3, 0.3))

  ggsave(plot, width = 170, height = 250, units = "mm", dpi = 300,
         file = here("Figures/Figure 2.eps"), device = cairo_ps)

```

# Figure 3 - Overlap in depth-enrichment of viral populations between sites

```{r save proportion of community abundance represented by depth-specific vOTUs (Hilly grassland)}

# save raw vOTU reads
  raw_reads <- 
    DNA_vOTU_reads %>% 
    select(NAME, contains("Sage")) %>% 
    column_to_rownames("NAME")

# create column metadata table
  metadata <- 
    tibble(SAMPLE = colnames(raw_reads)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly_grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry_Oak")) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ "SUBSURFACE",
                              grepl("100", SAMPLE) ~ "SUBSURFACE",
                              grepl("80", SAMPLE) ~ "SUBSURFACE",
                              grepl("60", SAMPLE) ~ "SUBSURFACE",
                              grepl("40", SAMPLE) ~ "SUBSURFACE",
                              grepl("20", SAMPLE) ~ "SURFACE")) %>% 
    mutate(SITE = factor(SITE, levels = c("Hilly_grassland", "Garry_Oak"))) %>% 
    mutate(DEPTH = factor(DEPTH, levels = c("SURFACE", "SUBSURFACE")))
    
# construct DESeqDataSet object
  ddsFullCountTable <- DESeqDataSetFromMatrix(
    countData = raw_reads, 
    colData = metadata,
    design = ~ SITE * DEPTH)
  
# run DESeq2
  dds <- DESeq(ddsFullCountTable, fitType = "local", quiet = T)
  # to probe specific coefficient (listed by resultsNames(dds))

# save log2 fold changes in vOTU abundance by depth (in Hilly grassland)
  res_grassland <- as_tibble(results(dds, name = "DEPTH_SUBSURFACE_vs_SURFACE"), rownames = "NAME")
  
# save subsurface-enriched vOTUs
  signif_subsurface <- 
    res_grassland %>% 
    filter(padj < 0.05) %>% 
    filter(log2FoldChange > 1) %>% 
    pull("NAME")
  
# save surface-enriched vOTUs
  signif_surface <- 
    res_grassland %>% 
    filter(padj < 0.05) %>% 
    filter(log2FoldChange < -1) %>% 
    pull("NAME")
  
# save proportion of community abundance represented by depth-specific vOTUs (Hilly grassland)
  depth_means_grassland <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    select(NAME, contains("Sage")) %>%
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "CPM") %>% 
    mutate(SIGNIF = ifelse(NAME %in% signif_subsurface, "SUBSURFACE",
                           ifelse(NAME %in% signif_surface, "SURFACE", "NS"))) %>% 
    mutate(SIGNIF = factor(SIGNIF, levels = c("SUBSURFACE", "SURFACE", "NS"))) %>% 
    group_by(SAMPLE, SIGNIF) %>% 
    summarise(AggRelAb = sum(CPM), .groups = "drop") %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                              grepl("100", SAMPLE) ~ 100,
                              grepl("80", SAMPLE) ~ 80,
                              grepl("60", SAMPLE) ~ 60,
                              grepl("40", SAMPLE) ~ 40,
                              grepl("20", SAMPLE) ~ 20)) %>% 
    group_by(SITE, DEPTH, SIGNIF) %>% 
    summarise(MeanRelAb = mean(AggRelAb), .groups = "drop") %>% 
    mutate(CONTRAST = "Effect of depth in Hilly grassland")

```

```{r save proportion of community abundance represented by depth-specific vOTUs (Garry Oak)}

# save raw vOTU reads
  raw_reads <- 
    DNA_vOTU_reads %>% 
    select(NAME, contains("Sage")) %>% 
    column_to_rownames("NAME")

# create column metadata table
  metadata <- 
    tibble(SAMPLE = colnames(raw_reads)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly_grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry_Oak")) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ "SUBSURFACE",
                              grepl("100", SAMPLE) ~ "SUBSURFACE",
                              grepl("80", SAMPLE) ~ "SUBSURFACE",
                              grepl("60", SAMPLE) ~ "SUBSURFACE",
                              grepl("40", SAMPLE) ~ "SUBSURFACE",
                              grepl("20", SAMPLE) ~ "SURFACE")) %>% 
    mutate(SITE = factor(SITE, levels = c("Garry_Oak", "Hilly_grassland"))) %>% 
    mutate(DEPTH = factor(DEPTH, levels = c("SURFACE", "SUBSURFACE")))
    
# construct DESeqDataSet object
  ddsFullCountTable <- DESeqDataSetFromMatrix(
    countData = raw_reads, 
    colData = metadata,
    design = ~ SITE * DEPTH)
  
# run DESeq2
  dds <- DESeq(ddsFullCountTable, fitType = "local", quiet = T)
  # to probe specific coefficient (listed by resultsNames(dds))

# save log2 fold changes in vOTU abundance by depth (in Garry Oak)
  res_oak <- as_tibble(results(dds, name = "DEPTH_SUBSURFACE_vs_SURFACE"), rownames = "NAME")
  
# save subsurface-enriched vOTUs
  signif_subsurface <- 
    res_oak %>% 
    filter(padj < 0.05) %>% 
    filter(log2FoldChange > 1) %>% 
    pull("NAME")
  
# save surface-enriched vOTUs
  signif_surface <- 
    res_oak %>% 
    filter(padj < 0.05) %>% 
    filter(log2FoldChange < -1) %>% 
    pull("NAME")
  
# save proportion of community abundance represented by depth-specific vOTUs (Garry Oak)
  depth_means_oak <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    select(NAME, contains("Sage")) %>%
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "CPM") %>%  
    mutate(SIGNIF = ifelse(NAME %in% signif_subsurface, "SUBSURFACE",
                           ifelse(NAME %in% signif_surface, "SURFACE", "NS"))) %>% 
    mutate(SIGNIF = factor(SIGNIF, levels = c("SUBSURFACE", "SURFACE", "NS"))) %>% 
    group_by(SAMPLE, SIGNIF) %>% 
    summarise(AggRelAb = sum(CPM), .groups = "drop") %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                              grepl("100", SAMPLE) ~ 100,
                              grepl("80", SAMPLE) ~ 80,
                              grepl("60", SAMPLE) ~ 60,
                              grepl("40", SAMPLE) ~ 40,
                              grepl("20", SAMPLE) ~ 20)) %>% 
    group_by(SITE, DEPTH, SIGNIF) %>% 
    summarise(MeanRelAb = mean(AggRelAb), .groups = "drop") %>% 
    mutate(CONTRAST = "Effect of depth in Garry Oak")

```

```{r Figure 3A - Relative abundance of depth-enriched viral populations}

# visualise the relative abundance of depth-enriched viral populations
  plot <- 
    rbind(depth_means_grassland, depth_means_oak) %>% 
      filter(!SIGNIF == "NS") %>% 
      mutate(PROP = MeanRelAb/1e6) %>% 
      mutate(CONTRAST = fct_recode(CONTRAST, 
                                   `Effect of depth in\nHilly grassland` = "Effect of depth in Hilly grassland",
                                   `Effect of depth in\nGarry Oak` = "Effect of depth in Garry Oak")) %>% 
      ggplot(aes(x = DEPTH, y = PROP, fill = SIGNIF)) +
        geom_area(colour = "white") +
        facet_grid(cols = vars(SITE),
                   rows = vars(CONTRAST)) +
        scale_fill_manual(breaks = c("SURFACE", "SUBSURFACE"),
                          labels = c("Surface-enriched", "Subsurface-enriched"),
                          values = c("#B5651D", "#2A0001")) +
        scale_x_continuous(limits = c(20, 120),
                           breaks = seq(20, 120, 20)) +
        scale_y_continuous(limits = c(0, 1),
                           breaks = seq(0, 1, 0.25)) +
        theme(legend.position = "bottom",
              text = element_text(size = 10)) +
        labs(x = "Soil depth (cm)",
             y = "Proportion of community abundance", 
             fill = "",
             title = "")

  A <- plot_grid(plot, labels = "A", label_size = 15)
  
  rm(list = c("dds", "ddsFullCountTable"))

```

```{r Figure 3B - Overlap in depth-enrichment of viral populations between sites}

# create a presence-absence matrix for upset plot
  df <- 
    bind_rows(list("Hilly grassland" = res_grassland, "Garry Oak" = res_oak), .id = "SITE") %>%
    mutate(ENRICHMENT = case_when(log2FoldChange < -1 & padj < 0.05 ~ "Surface-enriched",
                                  log2FoldChange > 1 & padj < 0.05 ~ "Subsurface-enriched",
                                  TRUE ~ "No enrichment")) %>% 
    filter(!ENRICHMENT == "No enrichment") %>% 
    select(SITE, NAME, ENRICHMENT) %>% 
    mutate(PRESENCE = 1) %>% 
    pivot_wider(names_from = "SITE", values_from = "PRESENCE", values_fill = 0)

# visualise the overlap in depth-enrichment of viral populations between sites
  plot <-
    upset(data = df, 
          intersect = c("Hilly grassland", "Garry Oak"),
          name = "Sample intersection",
          set_sizes = upset_set_size() +
            labs(y = "Total vOTUs"),
          base_annotations = list(
            "vOTUs in\nintersection" = intersection_size(text = list(size = 3)) +
              scale_y_continuous(limits = c(0, 4000))),
          annotations = list(
            "Enriched vOTUs" = list(
              aes = aes(x = intersection, fill = fct_rev(ENRICHMENT)),
              geom = list(
                geom_bar(stat = "count", position = "fill", na.rm = T),
                scale_fill_manual(breaks = c("Surface-enriched", "Subsurface-enriched"),
                                  values = c("#B5651D", "#2A0001")),
                scale_y_continuous(labels = scales::percent_format()),
                theme_bw(),
                theme(axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(),
                      #axis.title.x = element_blank(),
                      text = element_text(size = 10),
                      legend.position = "none"),
                labs(title = "")))),
          matrix = intersection_matrix(geom = geom_point(size = 2)),
          themes = upset_modify_themes(
            list(
              "vOTUs in\nintersection" = theme_bw() + 
                theme(axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(),
                      text = element_text(size = 10)),
              "intersections_matrix" = theme_bw() + 
                theme(axis.title.y = element_blank(),
                      axis.text.x = element_blank(),
                      axis.ticks.x = element_blank(), 
                      axis.line.x = element_blank(),
                      text = element_text(size = 10)),
              "overall_sizes" = theme_bw() +
                theme(axis.title.y = element_blank(),
                      axis.text.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      axis.line.y = element_blank(),
                      text = element_text(size = 10))
              )
            ),
          stripes = c("grey95", "white"),
          width_ratio = 0.5,
          height_ratio = 0.3,
          sort_intersections_by = c("degree", "cardinality"),
          sort_intersections = c("descending"),
          group_by = "degree",
          sort_sets = "ascending")
  
  B <- plot_grid(plot, labels = "B", label_size = 15)

  rm(list = c("res_grassland", "res_oak"))
  
```

```{r Figure 3 - Overlap in depth-enrichment of viral populations between sites}

  plot <- plot_grid(A, B, ncol = 1, rel_heights = c(1, 0.8))
  
  ggsave(plot, width = 120, height = 200, units = "mm", dpi = 300,
         file = here("Figures/Figure 3.eps"), device = cairo_ps,  fallback_resolution = 300)

```

# Figure 4 - Strain-level assembly of soil viral communities throughout soil depth

```{r Figure 4A - Distance decay relationship in consensus ANI}

# calculate vOTU CPM
  CPM <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME")

# extract vOTUs present in > 90% of samples
  prevalent_vOTUs <- 
    as_tibble(CPM, rownames = "NAME") %>% 
    pivot_longer(-NAME, names_to = "SAMPLE", values_to = "CPM") %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    group_by(NAME) %>% 
    summarise(OCCUPANCY = sum(CPM > 0)) %>% 
    filter(OCCUPANCY >= 24 * 0.90) %>% 
    pull(NAME)
  
# import inStrain output
  comparisons <- 
    as_tibble(read.delim(here("Data/instrainComparer_comparisonsTable.tsv")))

# calculate average consensus ANI for each pairwise depth comparison
  ANI_depth <- 
    comparisons %>% 
    filter(scaffold %in% prevalent_vOTUs & percent_genome_compared > 0.25) %>% # 74,752 (5,031 vOTUs)
    mutate(across(c(name1, name2), list(DEPTH = ~case_when(grepl("113", .) | grepl("115", .) ~ 115,
                                                           grepl("100", .) ~ 100,
                                                           grepl("80", .) ~ 80,
                                                           grepl("60", .) ~ 60,
                                                           grepl("40", .) ~ 40,
                                                           grepl("20", .) ~ 20),
                                        SITE = ~case_when(grepl("Sage1", .) ~ "Hilly grassland",
                                                          grepl("Sage2", .) ~ "Garry Oak")))) %>%
    filter(name1_SITE == name2_SITE) %>% 
    mutate(DEPTH_DIFFERENCE = abs(name1_DEPTH - name2_DEPTH)) %>% 
    group_by(scaffold) %>% 
    filter(n_distinct(DEPTH_DIFFERENCE) >= 3) %>% 
    ungroup() %>% 
    rename("NAME" = scaffold,
           "SITE" = name1_SITE) %>% 
    select(NAME, SITE, conANI, DEPTH_DIFFERENCE)

# extract vOTUs with significant changes in consensus ANI across soil depth
  ANI_correlations <- 
    ANI_depth %>% 
    nest(data = -NAME) %>% 
    mutate(cor_test = map(data, ~cor.test(.x$conANI, .x$DEPTH_DIFFERENCE, method = "pearson", alternative = "two.sided")),
           tidied = map(cor_test, tidy)) %>% 
    unnest(tidied) %>% 
    ungroup() %>% 
    mutate(P_ADJ = p.adjust(p.value, method = "holm")) %>%
    filter(P_ADJ < 0.05) %>% 
    rename("ESTIMATE" = estimate) %>% 
    select(NAME, ESTIMATE, P_ADJ)
  
# compute correlation coefficients for consensus ANI and difference in soil depth
  stat_correlation <- 
    inner_join(ANI_depth, ANI_correlations, by = "NAME") %>% 
    nest(data = everything()) %>%
    mutate(cor_test = map(data, ~cor.test(.x$conANI, .x$DEPTH_DIFFERENCE, method = "pearson")),
           lm_test = map(data, ~lm(.x$conANI ~ .x$DEPTH_DIFFERENCE)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3)))
  
# plot lines of consensus ANI over differences in soil depth
  plot <- 
    inner_join(ANI_depth, ANI_correlations, by = "NAME") %>%
      ggplot(aes(x = DEPTH_DIFFERENCE, y = conANI * 100)) +
        geom_line(aes(group = NAME), 
                  stat = "smooth", method = "lm", formula = "y ~ x", se = F, 
                  colour = "grey40", alpha = 0.3) +
        geom_line(stat = "smooth", method = "lm", formula = "y ~ x", se = F,
                  colour = "black", size = 1.5) +
        geom_text(data = stat_correlation, aes(label = label),
                  x = 10, y = 96.5, size = 3, hjust = 0, lineheight = 1) +
        scale_x_continuous(limits = c(0, 100),
                           breaks = seq(0, 100, 20)) +
        scale_y_continuous(limits = c(95, 100.6),
                           breaks = seq(95, 100, 1)) +
        labs(x = "Depth difference (cm)",
             y = "Consensus ANI (%)",
             title = "") +
        theme(legend.position = "none",
              text = element_text(size = 10))
  
  A <- plot_grid(plot, labels = "A", label_size = 15)
  
  rm(list = c("ANI_correlations", "ANI_depth", "comparisons"))

```

```{r Bootstrap average nucleotide diversity (pi) estimation}

# import local-scale (i.e., within each sample) micro diversity
  local_contig_microdiversity <- 
    as_tibble(read.delim(here("Data/local_contig_microdiversity.tsv")))

# create a function to bootstrap pi estimation
  iterate_micro_pi <- function(x){

    set.seed(x)
    
    # record smallest sample size for sampling
    sample_size <-
      local_contig_microdiversity %>%
      select(source, contig, pi) %>%
      rename("SAMPLE" = source,
             "NAME" = contig,
             "PI" = pi) %>%
      group_by(SAMPLE) %>%
      summarise(NUM_vOTUs = n_distinct(NAME), .groups = "drop") %>%
      summarise(MIN = min(NUM_vOTUs))
    
    # compute average pi value from sample
    pi_values <- 
      local_contig_microdiversity %>% 
      select(source, contig, pi) %>% 
      rename("SAMPLE" = source, 
             "NAME" = contig,
             "PI" = pi) %>% 
      mutate(SAMPLE = str_replace(SAMPLE, "_sorted", "")) %>% 
      group_by(SAMPLE) %>% 
      slice_sample(n = sample_size$MIN) %>% 
      group_by(SAMPLE) %>% 
      summarise(MEDIAN = median(PI)) %>% 
      arrange(SAMPLE) %>% 
      deframe()
    
    # output average pi value for each sample
    c(pi_values)
  
  }
  set.seed(123)

# bootstrap pi estimation, save results, and compute averages
  micro_pi <- 
    future_map_dfr(1:1000, iterate_micro_pi, .id = "SEED") %>% 
    pivot_longer(cols = -SEED, names_to = "SAMPLE", values_to = "PI") %>% 
    group_by(SAMPLE) %>% 
    summarise(across(PI, list(MEDIAN = median, 
                              MEAN = mean, 
                              t = ~qt(0.95, n()-1),
                              SE = ~sd(.x)/sqrt(n()))))

```

```{r Figure 4B - Viral macro diversity and micro diversity throughout the soil depth profiles}

# calculate vOTU CPM
  CPM <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    mutate(across(contains("Sage"), ~round(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME")

# generate phyloSeq vOTU matrix
  phyloseq <- otu_table(CPM, taxa_are_rows = T)

# create table of observed vOTU abundance and alpha diversity measures 
  alpha_diversity <-
    estimate_richness(phyloseq) %>%
    as_tibble(., rownames = "SAMPLE") %>%
    mutate(Pielou = Shannon / log(Observed)) %>%
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(DEPTH2 = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ "115 cm",
                             grepl("100", SAMPLE) ~ "100 cm",
                             grepl("80", SAMPLE) ~ "80 cm",
                             grepl("60", SAMPLE) ~ "60 cm",
                             grepl("40", SAMPLE) ~ "40 cm",
                             grepl("20", SAMPLE) ~ "20 cm")) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))
  
# create data frame of sample-averaged micro diversity estimates
  microdiversity <-
    micro_pi %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                             grepl("100", SAMPLE) ~ 100,
                             grepl("80", SAMPLE) ~ 80,
                             grepl("60", SAMPLE) ~ 60,
                             grepl("40", SAMPLE) ~ 40,
                             grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))

  # visualise viral macro diversity and micro diversity along the soil depth profile
  B <-
    inner_join(alpha_diversity, microdiversity, by = c("SAMPLE", "DEPTH", "SITE")) %>% 
      ggplot(aes(x = DEPTH)) +
        geom_smooth(aes(y = PI_MEAN, colour = "PI_MEAN"), fill = "#67a9cf", 
                    method = "loess", formula = "y ~ x", se = T, alpha = 0.2, size = 1) +
        geom_smooth(aes(y = (Shannon - 6.5)/4000, colour = "Shannon"), fill = "#ef8a62", 
                    method = "loess", formula = "y ~ x", se = T, alpha = 0.2, size = 1) +
        facet_wrap(~ SITE, nrow = 1) +
        scale_colour_manual(breaks = c("Shannon", "PI_MEAN"),
                            values = c("#ef8a62", "#67a9cf"),
                            labels = c("Macro diversity", "Micro diversity")) +
        scale_y_continuous(labels = scales::comma,
                           limits = c(0, 0.0005),
                           breaks = seq(0, 0.0005, 0.0001),
                           sec.axis = sec_axis(~(. * 4000) + 6.5, name = "Macro diversity (Shannon's H)")) +
        scale_x_continuous(limits = c(20, 120),
                           breaks = seq(20, 120, 20)) +
        guides(colour = guide_legend(override.aes = list(fill = NA))) +
        theme(legend.position = "right",
              text = element_text(size = 10)) +
        labs(x = "Soil depth (cm)",
             y = "Micro diversity (average \u03C0)",
             colour = "",
             title = "")
  
  # B <- plot_grid(plot, labels = "B", label_size = 15)

```

```{r Figure 4C - Correlation of macro diversity with micro diversity}

# calculate vOTU CPM
  CPM <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    mutate(across(contains("Sage"), ~round(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME")

# generate phyloSeq vOTU matrix
  phyloseq <- otu_table(CPM, taxa_are_rows = T)

# create table of observed vOTU abundance and alpha diversity measures 
  alpha_diversity <-
    estimate_richness(phyloseq) %>%
    as_tibble(., rownames = "SAMPLE") %>%
    mutate(Pielou = Shannon / log(Observed)) %>%
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(DEPTH2 = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ "115 cm",
                             grepl("100", SAMPLE) ~ "100 cm",
                             grepl("80", SAMPLE) ~ "80 cm",
                             grepl("60", SAMPLE) ~ "60 cm",
                             grepl("40", SAMPLE) ~ "40 cm",
                             grepl("20", SAMPLE) ~ "20 cm")) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))

# create data frame of sample-averaged micro diversity estimates
  microdiversity <-
    micro_pi %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                             grepl("100", SAMPLE) ~ 100,
                             grepl("80", SAMPLE) ~ 80,
                             grepl("60", SAMPLE) ~ 60,
                             grepl("40", SAMPLE) ~ 40,
                             grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(DEPTH2 = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ "115 cm",
                              grepl("100", SAMPLE) ~ "100 cm",
                              grepl("80", SAMPLE) ~ "80 cm",
                              grepl("60", SAMPLE) ~ "60 cm",
                              grepl("40", SAMPLE) ~ "40 cm",
                              grepl("20", SAMPLE) ~ "20 cm")) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))
  
# compute correlation coefficients for viral macro diversity and micro diversity
  stat_correlation <- 
    inner_join(alpha_diversity, microdiversity, by = c("SAMPLE", "DEPTH", "DEPTH2", "SITE")) %>%
    nest(data = -SITE) %>%
    mutate(cor_test = map(data, ~cor.test(.x$Shannon, .x$PI_MEDIAN, method = "pearson")),
           lm_test = map(data, ~lm(.x$PI_MEDIAN ~ .x$Shannon)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>% 
    filter(!lm_tidied.term == "(Intercept)") %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3), 
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, label) %>% 
    mutate(XPOS = c(NA),
           YPOS = c(NA))
  
# visualise correlation plot of viral macro diversity and viral micro diversity
  C <-
    inner_join(alpha_diversity, microdiversity, by = c("SAMPLE", "DEPTH", "DEPTH2", "SITE")) %>%
      ggplot(aes(x = Shannon, y = PI_MEDIAN)) +
        geom_point(aes(shape = SITE, fill = DEPTH2), 
                   size = 4, alpha = 0.8, stroke = 0) +
        facet_grid(cols = vars(SITE)) +
        scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                           labels = c("Hilly grassland", "Garry Oak"),
                           values = c(22, 24)) +
        scale_fill_manual(breaks = c("20 cm", "40 cm", "60 cm", "80 cm", "100 cm", "115 cm"), 
                          values = browns_palette) +
        scale_y_continuous(labels = scales::comma,
                           limits = c(0, 0.0005),
                           breaks = seq(0, 0.0005, 0.0001)) +
        scale_x_continuous(limits = c(7, 8.5),
                           breaks = seq(7, 8.5, 0.25)) +
        guides(fill = guide_legend(order = 2, override.aes = list(shape = 21)),
               shape = guide_legend(order = 1, override.aes = list(fill = "white", stroke = 1, size = 3))) +
        theme(legend.position = "right",
              text = element_text(size = 10)) +
        labs(x = "Macro diversity (Shannon's H)",
             y = "Micro diversity (average \u03C0)",
             fill = "Soil depth:", 
             shape = "Site:",
             title = "")
  
  # C <- plot_grid(plot, labels = "C", label_size = 15)
  
```

```{r Figure 4 - Strain-level assembly of soil viral communities throughout soil depth}

  bottom <- plot_grid(B, C, labels = c("B", "C"), label_size = 15, ncol = 1, align = "v", axis = "lr")

  plot <- plot_grid(A, bottom, ncol = 1, rel_heights = c(0.5, 1))

  ggsave(plot, width = 170, height = 200, units = "mm", dpi = 300,
         file = here("Figures/Figure 4.eps"), device = cairo_ps,  fallback_resolution = 300)

```

## Virus-host interactions were diverse with soil depth

# Supplementary Table S4 - Microbial OTU normalised coverage

```{r Supplementary Table S4 - Microbial OTU normalised coverage}

  OTU_coverage <- 
    as_tibble(read.csv(here("Data/OTU_normalised_coverage.csv")))
  # 1,447 OTUs
  # normalised coverage values
    # (raw coverage / sample read depth) * mean read depth across all samples
  
  write.csv(OTU_coverage, file = here("Tables/Table S4.csv"), row.names = F)

```

# Supplementary Table S5 - Raw read bacterial class taxonomy

```{r Supplementary Table S5 - Raw read bacterial class taxonomy}

  kraken_reads <- 
    as_tibble(read.csv(here("Data/kraken_class_reads.csv")))
    # class_reads %>% 
    # mutate(CLASS = case_when(CLASS == "Actinomycetes" ~ "Actinomycetia",
    #                          CLASS == "Alphaproteobacteria" | CLASS == "Alphaproteobacteria incertae sedis" ~ "AlphaPseudomonadota",
    #                          CLASS == "Betaproteobacteria" | CLASS == "Betaproteobacteria incertae sedis" ~ "BetaPseudomonadota",
    #                          CLASS == "Deltaproteobacteria" | CLASS == "Deltaproteobacteria incertae sedis" ~ "DeltaPseudomonadota",
    #                          CLASS == "Gammaproteobacteria" | CLASS == "Gammaproteobacteria incertae sedis" | CLASS == "unclassified Gammaproteobacteria" ~ "GammaPseudomonadota",
    #                          TRUE ~ CLASS)) %>% 
    # filter(!if_any(everything(), ~ is.na(.))) %>% 
    # pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "READS") %>% 
    # group_by(CLASS, SAMPLE) %>% 
    # summarise(READS = sum(READS), .groups = "drop") %>% 
    # group_by(SAMPLE) %>% 
    # mutate(REL_ABUND = READS/sum(READS)) %>% 
    # select(-READS) %>% 
    # pivot_wider(names_from = "SAMPLE", values_from = "REL_ABUND")

  # relative abundance values
    # reads / sample reads

  write.csv(class_abundance_table, file = here("Tables/Table S5.csv"), row.names = F)

```

# Figure 5 - Virus-host interactions throughout soil depth

```{r Figure 5A - Virus-host linkages}

# create a data frame of viral abundance by host class
  vOTUs <- 
    DNA_vOTU_reads %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.)))) %>%
    group_by(HOST_CLASS) %>% 
    summarise(across(contains("Sage"), sum)) %>% 
    pivot_longer(-HOST_CLASS, names_to = "SAMPLE", values_to = "PROP") %>% 
    group_by(HOST_CLASS) %>% 
    mutate(HOST_CLASS = case_when(is.na(HOST_CLASS) ~ "Undefined",
                                  sum(PROP) < 0.1 ~ "Other",
                                  TRUE ~ HOST_CLASS)) %>% 
    group_by(HOST_CLASS, SAMPLE) %>% 
    summarise(vOTU = sum(PROP), .groups = "drop")

# create a data frame of OTU abundance by class  
  OTUs <-
    OTU_coverage %>%
    mutate(across(contains("Sage"), ~(./sum(.)))) %>%
    group_by(CLASS) %>%
    summarise(across(contains("Sage"), sum)) %>%
    pivot_longer(-CLASS, names_to = "SAMPLE", values_to = "PROP") %>%
    mutate(CLASS = case_when(CLASS == "Actinomycetia" ~ CLASS,
                             CLASS == "AlphaPseudomonadota" ~ CLASS,
                             CLASS == "BetaPseudomonadota" ~ CLASS,
                             CLASS == "Clostridia" ~ CLASS,
                             CLASS == "DeltaPseudomonadota" ~ CLASS,
                             CLASS == "GammaPseudomonadota" ~ CLASS,
                             CLASS == "Vicinamibacteria" ~ CLASS,
                             CLASS == "Mollicutes" ~ CLASS,
                             CLASS == "Undefined" ~ CLASS,
                             TRUE ~ "Other")) %>%
    group_by(CLASS, SAMPLE) %>%
    summarise(OTU = sum(PROP), .groups = "drop")
  
# create a data frame of kraken reads abundance by class  
  kraken <- 
    kraken_reads %>% 
    mutate(across(contains("Sage"), ~(./sum(.)))) %>%
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "PROP") %>% 
    mutate(CLASS = case_when(CLASS == "Actinomycetia" ~ CLASS,
                             CLASS == "AlphaPseudomonadota" ~ CLASS,
                             CLASS == "BetaPseudomonadota" ~ CLASS,
                             CLASS == "Clostridia" ~ CLASS,
                             CLASS == "DeltaPseudomonadota" ~ CLASS,
                             CLASS == "GammaPseudomonadota" ~ CLASS,
                             CLASS == "Vicinamibacteria" ~ CLASS,
                             CLASS == "Mollicutes" ~ CLASS,
                             CLASS == "Undefined" ~ CLASS,
                             TRUE ~ "Other")) %>%
    group_by(CLASS, SAMPLE) %>%
    summarise(KRAKEN = sum(PROP), .groups = "drop") 
  
# create a data frame of abundance by phylum for vOTUs, OTUs, and MAGs
  df <-
    full_join(OTUs, vOTUs, by = c("SAMPLE", "CLASS" = "HOST_CLASS")) %>% 
    full_join(kraken, by = c("SAMPLE", "CLASS")) %>% 
    mutate(CLASS = factor(CLASS, levels = c("Other", "Mollicutes", "Vicinamibacteria", "Clostridia", "GammaPseudomonadota", "AlphaPseudomonadota", "BetaPseudomonadota", "DeltaPseudomonadota", "Actinomycetia"))) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    group_by(CLASS, DEPTH, SITE) %>% 
    summarise(across(c(OTU, vOTU, KRAKEN), ~mean(.)), .groups = "drop") %>% 
    replace_na(replace = list(OTU = 0, vOTU = 0, MAG = 0, KRAKEN = 0)) %>% 
    pivot_longer(c(OTU, vOTU, KRAKEN), names_to = "GENOME", values_to = "PROPORTION") %>% 
    mutate(GENOME = factor(GENOME, levels = c("vOTU", "OTU", "KRAKEN"))) %>% 
    mutate(GENOME = fct_recode(GENOME,
                               "vOTUs\n(n = 10,196)" = "vOTU",
                               "OTUs\n(n = 1,447)" = "OTU",
                               "Raw reads" = "KRAKEN")) %>% 
    filter(CLASS != "Undefined")
  
# visualise virus-host linkages
  plot <-
    ggplot(df, aes(x = DEPTH, y = PROPORTION, fill = CLASS)) +
      geom_area(colour = "white", position = "stack", outline.type = "upper", linewidth = 0.3) +
      facet_grid(cols = vars(SITE), 
                 rows = vars(GENOME)) +
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      scale_y_continuous(limits = c(0, 0.8),
                         breaks = seq(0, 0.8, 0.2)) +
      scale_fill_manual(breaks = c("Actinomycetia", "DeltaPseudomonadota", "BetaPseudomonadota", "AlphaPseudomonadota", "GammaPseudomonadota", "Clostridia", "Vicinamibacteria", "Mollicutes", "Other"),
                        labels = c("*Actinomycetia*", "*DeltaPseudomonadota*", "*BetaPseudomonadota*", "*AlphaPseudomonadota*", "*GammaPseudomonadota*", "*Clostridia*", "*Vicinamibacteria*", "*Mollicutes*", "Other"),
                        values = c(brewer.pal(8, "Dark2"), "grey80"),
                        drop = FALSE) +
      theme(text = element_text(size = 10),
            legend.text = element_markdown()) +
      guides(fill = guide_legend(reverse = T,
                                 override.aes = list(size = 0.5, colour = c(rep("white", 9))))) +
      labs(x = "Soil depth (cm)",
           y = "Mean proportional abundance",
           fill = "Bacterial class:",
           title = "")

  A <- plot_grid(plot, labels = "A", label_size = 15)
  
```

```{r Figure 5B - Incidence of lysogeny}

# create a data frame of temperate vOTU incidence
  df <- 
    DNA_vOTU_reads %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>% 
    select(NAME, LIFE_CYCLE, contains("Sage")) %>%
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "CPM") %>% 
    filter(CPM > 0) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    group_by(SAMPLE, DEPTH, SITE, LIFE_CYCLE) %>% 
    summarise(NUM_vOTUs = n(), .groups = "drop") %>% 
    group_by(SAMPLE, DEPTH, SITE) %>% 
    mutate(PROP_vOTUs = NUM_vOTUs/sum(NUM_vOTUs)) %>%
    filter(LIFE_CYCLE == "TEMPERATE")

# compute correlation coefficients for temperate vOTU incidence across soil depth
  stat_correlation <-
    df %>%
    nest(data = c(-SITE)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$DEPTH, .x$PROP_vOTUs, method = "pearson")),
           lm_test = map(data, ~lm(.x$PROP_vOTUs ~ .x$DEPTH)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    filter(cor_tidied.p.value < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, label) %>% 
    mutate(XPOS = c(30),
           YPOS = c(0.17))

# visualise the incidence of lysogeny
  plot <-
    ggplot(df, aes(x = DEPTH, y = PROP_vOTUs)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(data = filter(df, SITE %in% stat_correlation$SITE),
                  colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 3, lineheight = 1) +
      facet_grid(cols = vars(SITE)) +
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      scale_y_continuous(limits = c(0, 0.5),
                         breaks = seq(0, 0.5, 0.1)) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                           labels = c("Hilly grassland", "Garry Oak"),
                           values = c(15, 17),
                           guide = "none") +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)",
           y = "Incidence of lysogeny\n(Proportion of vOTUs)",
           title = "")

  B <- plot_grid(plot, labels = "B", label_size = 15)
  
```

```{r Figure 5C - Temperate viral abundance}

# create a data frame of temperate vOTU abundance
  df <- 
    DNA_vOTU_reads %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.)))) %>% 
    group_by(LIFE_CYCLE) %>% 
    summarise(across(contains("Sage"), ~sum(.))) %>% 
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "PROPORTION") %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    filter(LIFE_CYCLE == "TEMPERATE")

# compute correlation coefficients for temperate vOTU abundance across soil depth
  stat_correlation <-
    df %>%
    nest(data = c(-SITE)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$DEPTH, .x$PROPORTION, method = "pearson")),
           lm_test = map(data, ~lm(.x$PROPORTION ~ .x$DEPTH)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    filter(cor_tidied.p.value < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, label) %>% 
    mutate(XPOS = c(30),
           YPOS = c(0.17))
  
# visualise temperate viral abundance
  plot <- 
    ggplot(df, aes(x = DEPTH, y = PROPORTION)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(data = filter(df, SITE %in% stat_correlation$SITE),
                  colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 3, lineheight = 1) +
      facet_grid(cols = vars(SITE)) +
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      scale_y_continuous(limits = c(0, 0.5),
                         breaks = seq(0, 0.5, 0.1)) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                           labels = c("Hilly grassland", "Garry Oak"),
                           values = c(15, 17),
                           guide = "none") +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)",
           y = "Proportion of temperate\nviral abundance",
           title = "")
  
  C <- plot_grid(plot, labels = "C", label_size = 15)

```

```{r Figure 5D - Incidence of AMG carriage}

# create a data frame of incidence of vOTUs carrying AMGs
  df <- 
    DNA_vOTU_reads %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>% 
    mutate(AMG = case_when(NAME %in% AMG_annotations$NAME ~ "YES",
                           TRUE ~ "NO")) %>% 
    select(NAME, AMG, contains("Sage")) %>%
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "CPM") %>% 
    filter(CPM > 0) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    group_by(SAMPLE, DEPTH, SITE, AMG) %>% 
    summarise(NUM_vOTUs = n(), .groups = "drop") %>% 
    group_by(SAMPLE, DEPTH, SITE) %>% 
    mutate(PROP_vOTUs = NUM_vOTUs/sum(NUM_vOTUs)) %>%
    filter(AMG == "YES")

# compute correlation coefficients for incidence of vOTUs carrying AMGs across soil depth
  stat_correlation <-
    df %>%
    nest(data = c(-SITE)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$DEPTH, .x$PROP_vOTUs, method = "pearson")),
           lm_test = map(data, ~lm(.x$PROP_vOTUs ~ .x$DEPTH)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    filter(cor_tidied.p.value < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, label) %>% 
    mutate(XPOS = c(30),
           YPOS = c(0.009))

# visualise incidence of AMG carriage
  plot <-
    ggplot(df, aes(x = DEPTH, y = PROP_vOTUs)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(data = filter(df, SITE %in% stat_correlation$SITE),
                  colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 3, lineheight = 1) +
      facet_grid(cols = vars(SITE)) +
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      scale_y_continuous(limits = c(0, 0.04),
                         breaks = seq(0, 0.04, 0.01)) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                           labels = c("Hilly grassland", "Garry Oak"),
                           values = c(15, 17),
                           guide = "none") +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)",
           y = "Incidence of AMG carriage\n(Proportion of vOTUs)",
           title = "")

  D <- plot_grid(plot, labels = "D", label_size = 15)

```

```{r Figure 5E - AMG-carrying viral abundance}

# create a data frame of abundance of vOTUs carrying AMGs
  df <- 
    DNA_vOTU_reads %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.)))) %>% 
    filter(NAME %in% AMG_annotations$NAME) %>% 
    summarise(across(contains("Sage"), ~sum(.))) %>% 
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "PROPORTION") %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))

# compute correlation coefficients for abundance of vOTUs carrying AMGs across soil depth
  stat_correlation <-
    df %>%
    nest(data = c(-SITE)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$DEPTH, .x$PROPORTION, method = "pearson")),
           lm_test = map(data, ~lm(.x$PROPORTION ~ .x$DEPTH)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    filter(cor_tidied.p.value < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, label) %>% 
    mutate(XPOS = c(NA),
           YPOS = c(NA))
  
# visualise AMG-carrying viral abundance
  plot <-
    ggplot(df, aes(x = DEPTH, y = PROPORTION)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(data = filter(df, SITE %in% stat_correlation$SITE),
                  colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 3, lineheight = 1) +
      facet_grid(cols = vars(SITE)) +
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      scale_y_continuous(limits = c(0, 0.04),
                         breaks = seq(0, 0.04, 0.01)) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                           labels = c("Hilly grassland", "Garry Oak"),
                           values = c(15, 17),
                           guide = "none") +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)",
           y = "Proportion of AMG-carrying\nviral abundance",
           title = "")
  
  E <- plot_grid(plot, labels = "E", label_size = 15)

```

```{r Figure 5 - Virus-host interactions throughout soil depth}

  middle <- plot_grid(B, C, nrow = 1, rel_widths = c(1, 1))

  bottom <- plot_grid(D, E, nrow = 1, rel_widths = c(1, 1))

  plot <- plot_grid(A, middle, bottom, ncol = 1, rel_heights = c(0.5, 0.25, 0.25))

  ggsave(plot, width = 170, height = 250, units = "mm", dpi = 300,
         file = here("Figures/Figure 5.eps"), device = cairo_ps,  fallback_resolution = 300)

```

# Supplementary Figure S5 - Correlation of viral community and microbial community A structure and B diversity

```{r run mantel test}

# create matrix of OTU coverage values
  coverage <- 
    OTU_coverage %>% 
    select(NAME, contains("Sage")) %>% 
    column_to_rownames("NAME") %>% 
    t()

# compute Bray-Curtis dissimilarity on OTU coverage values
  dist_OTU <- 
    vegdist(coverage, method = "bray")
  
# calculate vOTU CPM
  CPM <- 
    DNA_vOTU_reads %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    mutate(across(contains("Sage"), ~sqrt(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME") %>% 
    t()

# compute Bray-Curtis dissimilarity on vOTU CPM values
  dist_vOTU <- 
    vegdist(CPM, method = "bray")
  
# mantel test
  mantel <- mantel(dist_OTU, dist_vOTU, permutations = 9999, method = "pearson")

```

```{r Supplementary Figure S5A - Correlation of viral community and microbial community structure}

# combine dissimilarity matrices for OTUs and vOTUs
  df <- 
    as_tibble(inner_join(dist2list(dist_OTU), dist2list(dist_vOTU), by = c("col", "row"))) %>% 
    rename("OTU" = value.x,
           "vOTU" = value.y) %>% 
    mutate(across(c(col, row), as.character)) %>%
    filter(col < row) %>%
    mutate(across(c(col, row), list(DEPTH = ~case_when(grepl("113", .) | grepl("115", .) ~ 115,
                                                       grepl("100", .) ~ 100,
                                                       grepl("80", .) ~ 80,
                                                       grepl("60", .) ~ 60,
                                                       grepl("40", .) ~ 40,
                                                       grepl("20", .) ~ 20),
                                    SITE = ~case_when(grepl("Sage1", .) ~ "Hilly grassland",
                                                      grepl("Sage2", .) ~ "Garry Oak")))) %>%
    filter(col_SITE == row_SITE) %>% 
    mutate(DEPTH_DIFFERENCE = abs(col_DEPTH - row_DEPTH))
  
# compute correlation coefficients for vOTU and OTU abundances
  stat_correlation <- 
    df %>% 
    nest(data = -col_SITE) %>%
    mutate(cor_test = map(data, ~cor.test(.x$OTU, .x$vOTU, method = "pearson")),
           lm_test = map(data, ~lm(.x$vOTU ~ .x$OTU)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    mutate(XPOS = c(0.74, 0.68),
           YPOS = c(0.18, 0.2))
  
# visualise correlations of vOTU and OTU community structures
  plot <-
    ggplot(df, aes(x = OTU, y = vOTU)) +
      geom_point(aes(shape = col_SITE), 
                 size = 2, alpha = 0.6, color = "gray25", stroke = 0) +
      geom_smooth(colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation,
                aes(x = XPOS, y = YPOS, label = label),
                size = 3, hjust = 0, lineheight = 1) +
      facet_grid(cols = vars(col_SITE)) +
      coord_fixed() +
      scale_x_continuous(limits = c(0, 1.0),
                         breaks = seq(0, 1.0, 0.2)) +
      scale_y_continuous(limits = c(0, 0.6),
                         breaks = seq(0, 0.6, 0.1)) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                         values = c(15, 17),
                         guide = "none") +
      theme(text = element_text(size = 10)) +
      labs(x = "OTU community dissimilarity",
           y = "vOTU community dissimilarity")
  
  A <- plot_grid(plot, labels = "A", label_size = 15)

```

```{r Supplementary Figure S5B - Correlation of viral community and microbial community diversity}

# estimate beta diversity of viral communities
  viral_diversity <- 
    DNA_vOTU_reads %>%
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.))*1e6)) %>%
    mutate(across(contains("Sage"), ~round(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME") %>%
    otu_table(., taxa_are_rows = T) %>%
    estimate_richness(., measures = c("Shannon")) %>%
    as_tibble(., rownames = "SAMPLE")

# estimate beta diversity of microbial communities
  microbial_diversity <-
    OTU_coverage %>%
    mutate(across(contains("Sage"), ~round(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME") %>%
    otu_table(., taxa_are_rows = T) %>%
    estimate_richness(., measures = c("Shannon")) %>%
    as_tibble(., rownames = "SAMPLE")

# create a data frame of community beta diversities
  df <- 
    inner_join(viral_diversity, microbial_diversity, by = "SAMPLE") %>% 
    rename("SHANNON_VIRAL" = Shannon.x,
           "SHANNON_MICROBIAL" = Shannon.y) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(DEPTH2 = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ "115 cm",
                             grepl("100", SAMPLE) ~ "100 cm",
                             grepl("80", SAMPLE) ~ "80 cm",
                             grepl("60", SAMPLE) ~ "60 cm",
                             grepl("40", SAMPLE) ~ "40 cm",
                             grepl("20", SAMPLE) ~ "20 cm")) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))
  
# compute correlation coefficients for viral and bacterial community beta diversities
  stat_correlation <- 
    df %>%
    nest(data = everything()) %>% 
    mutate(cor_test = map(data, ~cor.test(.x$SHANNON_MICROBIAL, .x$SHANNON_VIRAL, method = "pearson")),
           lm_test = map(data, ~lm(.x$SHANNON_VIRAL ~ .x$SHANNON_MICROBIAL)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>% 
    filter(!lm_tidied.term == "(Intercept)") %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3), 
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(label) %>% 
    mutate(XPOS = 5.9,
           YPOS = 8.05)
  
# visualise correlation of viral community and microbial community divserity
  plot <-
    ggplot(df, aes(x = SHANNON_MICROBIAL, y = SHANNON_VIRAL)) +
      geom_point(aes(shape = SITE, fill = DEPTH2), 
                 size = 3, alpha = 0.8, stroke = 0) +
      geom_smooth(colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.2) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 3, lineheight = 1) +
      coord_fixed() +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                           labels = c("Hilly grassland", "Garry Oak"),
                           values = c(22, 24)) +
      scale_fill_manual(breaks = c("20 cm", "40 cm", "60 cm", "80 cm", "100 cm", "115 cm"), 
                        values = browns_palette) +
      scale_x_continuous(limits = c(5.8, 6.6),
                         breaks = seq(5.8, 6.6, 0.2)) +
      scale_y_continuous(limits = c(7.2, 8.3),
                         breaks = seq(7.2, 8.3, 0.2)) +
      guides(fill = guide_legend(override.aes = list(shape = 21)),
             shape = guide_legend(override.aes = list(fill = "white", stroke = 1, size = 3))) +
      theme(legend.position = "right",
            text = element_text(size = 10)) +
      labs(x = "Microbial community diversity (Shannon's H)",
           y = "Viral community diversity (Shannon's H)",
           fill = "Soil depth:", 
           shape = "Site:",
           title = "")
  
  B <- plot_grid(plot, labels = "B", label_size = 15)

```

```{r Supplementary Figure S5 - Correlation of viral community and microbial community A structure and B diversity}

  plot <- plot_grid(A, B, ncol = 1)

  ggsave(plot, width = 170, height = 180, units = "mm", dpi = 300,
         file = here("Figures/Figure S5.pdf"), device = cairo_pdf)

```

# Supplementary Figure S6 - Correlation of viral abundances and host abundances

```{r Supplementary Figure S6 - Correlation of viral abundances and host abundances}

# create a data frame of viral abundance by host class
  viral_abundance <- 
    DNA_vOTU_reads %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.)))) %>%
    group_by(HOST_CLASS) %>% 
    summarise(across(contains("Sage"), ~sum(.))) %>% 
    filter(!is.na(HOST_CLASS) & !HOST_CLASS == "Unassigned") %>% 
    pivot_longer(contains("Sage"), names_to = "SAMPLE", values_to = "VIRAL_ABUNDANCE")

# create a data frame of microbial abundance by class
  bacterial_abundance <- 
    OTU_coverage %>% 
    mutate(across(contains("Sage"), ~(./sum(.)))) %>%
    group_by(CLASS) %>% 
    summarise(across(contains("Sage"), ~sum(.))) %>% 
    filter(!CLASS == "Undefined") %>% 
    pivot_longer(contains("Sage"), names_to = "SAMPLE", values_to = "BACTERIAL_ABUNDANCE")
  
# create a data frame of raw read abundance by class
  kraken_abundance <- 
    kraken_reads %>% 
    mutate(across(contains("Sage"), ~(./sum(.)))) %>%
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "PROP") %>% 
    mutate(CLASS = case_when(CLASS == "Actinomycetia" ~ CLASS,
                             CLASS == "AlphaPseudomonadota" ~ CLASS,
                             CLASS == "BetaPseudomonadota" ~ CLASS,
                             CLASS == "Clostridia" ~ CLASS,
                             CLASS == "DeltaPseudomonadota" ~ CLASS,
                             CLASS == "GammaPseudomonadota" ~ CLASS,
                             CLASS == "Vicinamibacteria" ~ CLASS,
                             CLASS == "Mollicutes" ~ CLASS,
                             CLASS == "Undefined" ~ CLASS,
                             TRUE ~ "Other")) %>%
    group_by(CLASS, SAMPLE) %>%
    summarise(KRAKEN_ABUNDANCE = sum(PROP), .groups = "drop")
  
# create a data frame of viral and microbial abundances by class
  df <- 
    inner_join(viral_abundance, bacterial_abundance, by = c("HOST_CLASS" = "CLASS", "SAMPLE")) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                             grepl("100", SAMPLE) ~ 100,
                             grepl("80", SAMPLE) ~ 80,
                             grepl("60", SAMPLE) ~ 60,
                             grepl("40", SAMPLE) ~ 40,
                             grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    mutate(GROUP = paste(SITE, ": *", HOST_CLASS, "*", sep = "")) %>% 
    mutate(GROUP = factor(GROUP, levels = c("Garry Oak: *Actinomycetia*", "Hilly grassland: *Actinomycetia*", "Garry Oak: *AlphaPseudomonadota*", "Hilly grassland: *AlphaPseudomonadota*", "Garry Oak: *Clostridia*", "Hilly grassland: *Clostridia*", "Garry Oak: *GammaPseudomonadota*", "Hilly grassland: *GammaPseudomonadota*")))
  
# compute correlation coefficients for viral and bacterial abundance by phylum
  stat_correlation <-
    df %>%
    nest(data = c(-GROUP)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$BACTERIAL_ABUNDANCE, .x$VIRAL_ABUNDANCE, method = "pearson")),
           lm_test = map(data, ~lm(.x$BACTERIAL_ABUNDANCE ~ .x$VIRAL_ABUNDANCE)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    mutate(cor_tidied.p.value_adjusted = p.adjust(cor_tidied.p.value, method = "holm")) %>% 
    filter(cor_tidied.p.value_adjusted < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value_adjusted, 3))) %>% 
    select(GROUP, label) %>% 
    mutate(XPOS = c(0.32, 0.02),
           YPOS = c(0.28, 0.17))
  
# visualise correlations of viral abundances and host abundances
  plot <-
    df %>% 
    filter(HOST_CLASS %in% c("Actinomycetia", "AlphaPseudomonadota")) %>% 
    ggplot(., aes(x = VIRAL_ABUNDANCE, y = BACTERIAL_ABUNDANCE)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(data = filter(df, GROUP %in% stat_correlation$GROUP),
                  colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 2.5, lineheight = 1) +
      facet_wrap(~ GROUP, scales = "free", nrow = 4, ncol = 2) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                         labels = c("Hilly grassland", "Garry Oak"), 
                         values = c(15, 17),
                         guide = "none") + 
      theme(text = element_text(size = 10), strip.text = element_markdown()) +
      labs(x = "Proportional viral abundance", 
           y = "Proportional host abundance",
           title = "")
  
  A <- plot_grid(plot, labels = "A", label_size = 15)
  
# create a data frame of viral and microbial abundances by class
  df <- 
    inner_join(viral_abundance, kraken_abundance, by = c("HOST_CLASS" = "CLASS", "SAMPLE")) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                             grepl("100", SAMPLE) ~ 100,
                             grepl("80", SAMPLE) ~ 80,
                             grepl("60", SAMPLE) ~ 60,
                             grepl("40", SAMPLE) ~ 40,
                             grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    mutate(GROUP = paste(SITE, ": *", HOST_CLASS, "*", sep = "")) %>% 
    mutate(GROUP = factor(GROUP, levels = c("Garry Oak: *Actinomycetia*", "Hilly grassland: *Actinomycetia*", "Garry Oak: *AlphaPseudomonadota*", "Hilly grassland: *AlphaPseudomonadota*", "Garry Oak: *BetaPseudomonadota*", "Hilly grassland: *BetaPseudomonadota*", "Garry Oak: *Clostridia*", "Hilly grassland: *Clostridia*", "Garry Oak: *DeltaPseudomonadota*", "Hilly grassland: *DeltaPseudomonadota*", "Garry Oak: *GammaPseudomonadota*", "Hilly grassland: *GammaPseudomonadota*", "Garry Oak: *Mollicutes*", "Hilly grassland: *Mollicutes*", "Garry Oak: *Vicinamibacteria*", "Hilly grassland: *Vicinamibacteria*")))
  
# compute correlation coefficients for viral and bacterial abundance by phylum
  stat_correlation <-
    df %>%
    nest(data = c(-GROUP)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$KRAKEN_ABUNDANCE, .x$VIRAL_ABUNDANCE, method = "pearson")),
           lm_test = map(data, ~lm(.x$KRAKEN_ABUNDANCE ~ .x$VIRAL_ABUNDANCE)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    mutate(cor_tidied.p.value_adjusted = p.adjust(cor_tidied.p.value, method = "holm")) %>% 
    filter(cor_tidied.p.value_adjusted < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value_adjusted, 3))) %>% 
    select(GROUP, label) %>% 
    mutate(XPOS = c(0.32, 0.022, 0.03, 0.055, 0.0085),
           YPOS = c(0.37, 0.17, 0.0043, 4.1e-04, 0.0021))
  
# visualise correlations of viral abundances and host abundances
  plot <-
    df %>% 
    filter(HOST_CLASS %in% c("Actinomycetia", "AlphaPseudomonadota", "Clostridia", "DeltaPseudomonadota", "Vicinamibacteria")) %>% 
    ggplot(., aes(x = VIRAL_ABUNDANCE, y = KRAKEN_ABUNDANCE)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(data = filter(df, GROUP %in% stat_correlation$GROUP),
                  colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 2.5, lineheight = 1) +
      facet_wrap(~ GROUP, scales = "free", nrow = 5, ncol = 2) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                         labels = c("Hilly grassland", "Garry Oak"), 
                         values = c(15, 17),
                         guide = "none") + 
      theme(text = element_text(size = 10), strip.text = element_markdown()) +
      labs(x = "Proportional viral abundance", 
           y = "Proportional host abundance",
           title = "")
  
  B <- plot_grid(plot, labels = "B", label_size = 15)
  
  plot <- plot_grid(A, B, ncol = 1, rel_heights = c(2.2, 5))
  
  ggsave(plot, width = 170, height = 250, units = "mm", dpi = 300,
         file = here("Figures/Figure S6.pdf"), device = cairo_pdf)

```

# Supplementary Table S6 - Annotations of viral-encoded auxiliary metabolic genes

```{r Supplementary Table S6 - Annotations of viral-encoded auxiliary metabolic genes}

  AMG_annotations <- 
    as_tibble(read.csv(file = here("Data/AMG_annotations.csv")))

  write.csv(AMG_annotations, file = here("Tables/Table S6.csv"), row.names = F)

```

# Table 1 - Summary of viral-encoded CAZymes

```{r Table 1 - Summary of viral-encoded CAZymes}

  # made in excel, found at /Tables/Table 1.xlsx

```

# Supplementary Figure S7 - Relative abundance of viruses carrying carbohydrate-active enzymes

```{r Supplementary Figure S7 - Relative abundance of viruses carrying carbohydrate-active enzymes}

# create a data frame of relative abundance of CAZyme-carrying vOTUs
  df <- 
    DNA_vOTU_reads %>% 
    select(NAME, LENGTH, contains("Sage")) %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.)))) %>% 
    filter(NAME %in% filter(AMG_annotations, !cazy_hits == "")$NAME) %>% 
    summarise(across(contains("Sage"), ~sum(.))) %>% 
    pivot_longer(cols = contains("Sage"), names_to = "SAMPLE", values_to = "PROPORTION") %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    select(SAMPLE, DEPTH, SITE, PROPORTION)
    
# compute correlation coefficients for relative abundance of CAZyme-carrying vOTUs across soil depth
  stat_correlation <-
    df %>%
    nest(data = c(-SITE)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$PROPORTION, .x$DEPTH, method = "pearson")),
           lm_test = map(data, ~lm(.x$PROPORTION ~ .x$DEPTH)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    filter(cor_tidied.p.value < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, label) %>% 
    mutate(XPOS = NA,
           YPOS = NA)

# plot relative abundance of viruses carrying CAZyme AMGs
  plot <- 
    ggplot(df, aes(x = DEPTH, y = PROPORTION)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      # geom_smooth(data = filter(df, SITE %in% stat_correlation$SITE),
      #             colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      # geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
      #           hjust = 0, size = 3, lineheight = 1) +
      facet_grid(cols = vars(SITE)) +
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      scale_y_continuous(limits = c(0, 0.007),
                         breaks = seq(0, 0.007, 0.001)) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                           labels = c("Hilly grassland", "Garry Oak"),
                           values = c(15, 17),
                           guide = "none") +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)",
           y = "Proportional viral abundance",
           title = "")
    
  ggsave(plot, width = 170, height = 100, units = "mm", dpi = 300,
         file = here("Figures/Figure S7.pdf"), device = cairo_pdf)

```

## Virus-host antagonistic co-evolution was dynamic throughout the soil depth gradient

# Supplementary Table S7 - Microbial MAG normalised coverage

```{r Supplementary Table S7 - Microbial MAG normalised coverage}

  MAG_coverage <- 
    as_tibble(read.csv(here("Data/MAG_normalised_coverage.csv")))
  # 285 MAGs
  # normalised coverage values
    # (raw coverage / sample read depth) * mean read depth across all samples

  write.csv(MAG_coverage, file = here("Tables/Table S7.csv"), row.names = F)

```

# Figure 6 - Virus-host antagonistic co-evolution across soil depth profiles

```{r Figure 6A - Anti-phage system detection}

# import scaffold bin index
  scaffold_bin_index <- 
    as_tibble(read.csv(here("Data/scaffold_bin_index.csv"), header = T))

# import defense-finder output and filter for complete system detection
  whole_systems <- 
    as_tibble(read.delim(here("Data/defense_finder_genes.tsv"))) %>% 
    filter(sys_wholeness == 1)

# link anti-phage system detection to MAG bins
  anti_phage_systems <- 
    as_tibble(read.delim(here("Data/defense_finder_systems.tsv"))) %>% 
    mutate("SCAFFOLD" = word(sys_id, start = 1, end = 6, sep = "_")) %>% 
    left_join(scaffold_bin_index, by = "SCAFFOLD") %>% 
    filter(sys_id %in% whole_systems$sys_id) %>% 
    select(BIN, "TYPE" = type, "SUBTYPE" = subtype)
  
# create a data frame of relative abundance of MAGs carrying anti-phage defense systems
  df <-
    MAG_coverage %>% 
    pivot_longer(-BIN, names_to = "SAMPLE", values_to = "COVERAGE") %>% 
    group_by(SAMPLE) %>% 
    mutate(TOTAL_COVERAGE = sum(COVERAGE)) %>% 
    group_by(BIN) %>% 
    mutate(PROP_COVERAGE = COVERAGE/TOTAL_COVERAGE) %>% 
    left_join(anti_phage_systems, by = "BIN", relationship = "many-to-many") %>% 
    mutate(TYPE = case_when(is.na(TYPE) ~ "None",
                            TRUE ~ TYPE)) %>% 
    group_by(SAMPLE, TYPE) %>% 
    summarise(PROP_COVERAGE = sum(PROP_COVERAGE), .groups = "drop") %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                             grepl("100", SAMPLE) ~ 100,
                             grepl("80", SAMPLE) ~ 80,
                             grepl("60", SAMPLE) ~ 60,
                             grepl("40", SAMPLE) ~ 40,
                             grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    group_by(DEPTH, SITE, TYPE) %>% 
    summarise(PROP_COVERAGE = mean(PROP_COVERAGE), .groups = "drop") %>% 
    mutate(TYPE = case_when(TYPE == "None" | TYPE == "RM" | TYPE == "AbiEii" | TYPE == "SoFIC" | TYPE == "CBASS" | TYPE == "Wadjet" | TYPE == "Cas" | TYPE == "PsyrTA" ~ TYPE,
                              TRUE ~ "Other")) %>%
    group_by(DEPTH, SITE, TYPE) %>% 
    summarise(PROP_COVERAGE = sum(PROP_COVERAGE), .groups = "drop")

# plot anti-phage system detection
  A <-
    df %>% 
      ggplot(aes(x = DEPTH, y = PROP_COVERAGE, fill = factor(TYPE, levels = c("None", "Other", "PsyrTA", "Cas", "Wadjet", "CBASS", "SoFIC", "AbiEii", "RM")))) +
        geom_area(colour = "white", position = "fill") +
        facet_grid(cols = vars(SITE)) +
        scale_x_continuous(limits = c(20, 120),
                           breaks = seq(20, 120, 20)) +
        scale_fill_manual(breaks = c("None", "Other", "PsyrTA", "Cas", "Wadjet", "CBASS", "SoFIC", "AbiEii", "RM"),
                          values = c(alpha("white", 0), "grey80", brewer.pal(7, "Dark2"))) +
        guides(fill = guide_legend(override.aes = list(colour = c("black", rep("white", 8))))) +
        theme(text = element_text(size = 10)) +
        labs(x = "Soil depth (cm)",
             y = "Mean proportion of MAG abundance", 
             fill = "Anti-phage system:",
             title = "")
  
  # A <- plot_grid(plot, labels = "A", label_size = 15)

```

```{r Figure 6B - Diversity of the anti-phage system repertoire}

# calculate relative MAG abundance by anti-phage system
  coverage <-
    MAG_coverage %>% 
    pivot_longer(-BIN, names_to = "SAMPLE", values_to = "COVERAGE") %>% 
    group_by(SAMPLE) %>% 
    mutate(TOTAL_COVERAGE = sum(COVERAGE)) %>% 
    group_by(BIN) %>% 
    mutate(REL_COVERAGE = (COVERAGE/TOTAL_COVERAGE)*1e6) %>% 
    left_join(anti_phage_systems, by = "BIN") %>% 
    mutate(TYPE = case_when(is.na(TYPE) ~ "None",
                            TRUE ~ TYPE)) %>% 
    group_by(SAMPLE, TYPE) %>% 
    summarise(REL_COVERAGE = round(sum(REL_COVERAGE)), .groups = "drop") %>% 
    pivot_wider(names_from = SAMPLE, values_from = REL_COVERAGE) %>% 
    column_to_rownames("TYPE")

# generate phyloSeq vOTU matrix
  phyloseq <- otu_table(coverage, taxa_are_rows = T)

# create a data frame of anti-phage system diversity 
  df <-
    estimate_richness(phyloseq, measures = c("Shannon")) %>%
    as_tibble(., rownames = "SAMPLE") %>%
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(DEPTH2 = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ "115 cm",
                             grepl("100", SAMPLE) ~ "100 cm",
                             grepl("80", SAMPLE) ~ "80 cm",
                             grepl("60", SAMPLE) ~ "60 cm",
                             grepl("40", SAMPLE) ~ "40 cm",
                             grepl("20", SAMPLE) ~ "20 cm")) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))
  
# compute correlation coefficients for anti-phage system diversity across soil depth
  stat_correlation <-
    df %>%
    nest(data = -SITE) %>%
    mutate(cor_test = map(data, ~cor.test(.x$Shannon, .x$DEPTH, method = "pearson")),
           lm_test = map(data, ~lm(.x$Shannon ~ .x$DEPTH)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    filter(cor_tidied.p.value < 0.05) %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, label) %>% 
    mutate(XPOS = c(30, 30),
           YPOS = c(2.32, 2.3))
  
# plot diversity of the anti-phage system repertoire
  B <- 
    ggplot(df, aes(x = DEPTH, y = Shannon)) +
      geom_point(aes(shape = SITE),
                 size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 3, lineheight = 1) +
      facet_grid(cols = vars(SITE)) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                         labels = c("Hilly grassland", "Garry Oak"), 
                         values = c(15, 17),
                         guide = "none") + 
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      scale_y_continuous(limits = c(1.8, 2.42),
                         breaks = seq(1.8, 2.4, 0.2)) +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)", 
           y = "Diversity of anti-phage systems\n(Shannon's H)",
           title = "")
  
  # B <- plot_grid(plot, labels = "B", label_size = 15)

```

```{r Figure 6C - Viruses under positive selection}

# create a data frame of number of vOTUs under positive selection using micro diversity output from metapop
  df <-
    as_tibble(read.delim(here("Data/local_gene_microdiversity.tsv"))) %>%
    filter(!is.na(pNpS_ratio) & !is.infinite(pNpS_ratio)) %>%
    mutate(SAMPLE = str_replace(source, "_sorted", "")) %>%
    select(SAMPLE, "NAME" = contig, "GENE" = contig_gene, pNpS_ratio) %>%
    filter(pNpS_ratio > 1) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                             grepl("100", SAMPLE) ~ 100,
                             grepl("80", SAMPLE) ~ 80,
                             grepl("60", SAMPLE) ~ 60,
                             grepl("40", SAMPLE) ~ 40,
                             grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    group_by(SAMPLE, SITE, DEPTH) %>% 
    summarise(NUM_vOTUs = n_distinct(NAME), .groups = "drop")
  
# compute correlation coefficients for number of vOTUs under positive selection across soil depth
  stat_correlation <-
    df %>%
    nest(data = c(-SITE)) %>%
    mutate(cor_test = map(data, ~cor.test(.x$DEPTH, .x$NUM_vOTUs, method = "pearson")),
           lm_test = map(data, ~lm(.x$NUM_vOTUs ~ .x$DEPTH)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>%
    filter(!lm_tidied.term == "(Intercept)") %>%
    filter(cor_tidied.p.value < 0.05) %>%
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3),
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(SITE, label) %>% 
    mutate(XPOS = c(30),
           YPOS = c(55))

# plot viruses under positive selection
  C <-
    ggplot(df, aes(x = DEPTH, y = NUM_vOTUs)) +
      geom_point(aes(shape = SITE),
                 colour = "grey25", size = 2, alpha = 0.6, stroke = 0) +
      geom_smooth(data = filter(df, SITE %in% stat_correlation$SITE),
                  colour = "black", method = "lm", formula = "y ~ x", se = T, alpha = 0.4) +
      geom_text(data = stat_correlation, aes(x = XPOS, y = YPOS, label = label),
                hjust = 0, size = 3, lineheight = 1) +
      facet_grid(cols = vars(SITE)) +
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      scale_y_continuous(limits = c(0, 75),
                         breaks = seq(0, 75, 25)) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                           labels = c("Hilly grassland", "Garry Oak"),
                           values = c(15, 17),
                           guide = "none") +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)",
           y = "Number of vOTUs with\n\u2265 1 gene under selection",
           title = "")

  # C <- plot_grid(plot, labels = "C", label_size = 15)
    
```

```{r Figure 6 - Virus-host antagonistic co-evolution}

  plot <- plot_grid(A, B, C, ncol = 1, labels = c("A", "B", "C"), label_size = 15, rel_heights = c(1, 0.8, 0.8), align = "v", axis = "lr")

  ggsave(plot, width = 170, height = 200, units = "mm", dpi = 300,
         file = here("Figures/Figure 6.eps"), device = cairo_ps,  fallback_resolution = 300)

```

# Supplementary Table S8 - Annotations of viral genes under positive selection

```{r Supplementary Table S8 - Annotations of viral genes under positive selection}

# create a data frame of viral gene annotations
  curated_gene_annotations <- 
    DNA_vOTU_gene_annotations %>% 
    mutate(GENE = paste(NAME, gene_position, sep = "_")) %>% 
    select(GENE, LOCUS, contains("hit"), contains("ANNOTATION"), amg_flags)

# create a data frame of viral gene annotations for viral genes under positive selection
  pos_selection <-
    as_tibble(read.delim(here("Data/local_gene_microdiversity.tsv"))) %>%
    filter(!is.na(pNpS_ratio) & !is.infinite(pNpS_ratio)) %>%
    mutate(SAMPLE = str_replace(source, "_sorted", "")) %>%
    select(SAMPLE, "NAME" = contig, "GENE" = contig_gene, pNpS_ratio) %>%
    filter(pNpS_ratio > 1) %>%
    left_join(curated_gene_annotations, by = "GENE") %>%
    select(SAMPLE, NAME, LOCUS, pNpS_ratio, kegg_hit:amg_flags)

  write.csv(pos_selection, file = here("Tables/Table S8.csv"), row.names = F)

```

# Supplementary Figure S8 - Genome maps of high-quality viral genomes carrying ABC transporters under positive selection

```{r Supplementary Figure S8 - Genome maps of high-quality viral genomes carrying ABC transporters under positive selection}

# import viral genes with hits to lysogenic PFAM modules
  lysogeny_genes <- 
    as_tibble(read.delim(here("Data/lysogeny_gene_hits.tsv")))

# create a data frame of functional gene annotations for high-quality viral genomes carrying ABC transporters under positive selection
  df <-
    DNA_vOTU_gene_annotations %>% 
    filter(NAME == "DNA_vOTU_5734" | NAME == "DNA_vOTU_6575" | NAME == "DNA_vOTU_2761" | NAME == "DNA_vOTU_5919" | NAME == "DNA_vOTU_707") %>% 
    mutate(across(contains("position"), ~.x/1000)) %>% 
    mutate(FUNCTION = case_when(LOCUS %in% AMG_annotations$LOCUS ~ "AMG", # putative AMGs
                                # grepl("V", amg_flags) ~ "Structural/Replication",
                                # grepl("P", amg_flags) ~ "Peptidase",
                                # grepl("lysin", vogdb_hits) ~ "Lysis",
                                LOCUS == "SAGEDEP_086795" ~ "ABC transporter", # "ccmA",
                                LOCUS == "SAGEDEP_102900" ~ "ABC transporter", # "glycerol-3-phosphate transport system",
                                LOCUS == "SAGEDEP_048611" ~ "ABC transporter", # "MacAB",
                                LOCUS == "SAGEDEP_091261" ~ "ABC transporter", # "ABC-type nitrate/sulfonate/bicarbonate transport system",
                                LOCUS == "SAGEDEP_012472" ~ "ABC transporter", # "ABC-type Fe3+/spermidine/putrescine transport system",
                                LOCUS %in% lysogeny_genes$LOCUS ~ "Lysogeny", # lysogeny genes
                                grepl("phrog", PHROG) ~ "PHROG", # hits to PHROG database
                                TRUE ~ "CDS")) %>% 
    # group_by(FUNCTION) %>% count()
    select(NAME, LOCUS, FUNCTION, start_position, end_position, strandedness)
  
# plot genome map for DNA_vOTU_5734
  plot <-
    ggplot(filter(df, NAME == "DNA_vOTU_5734"), aes(xmin = start_position, xmax = end_position, y = NAME, forward = strandedness, fill = FUNCTION)) +
      geom_gene_arrow() +
      scale_fill_manual(values = c("yellow", "#D95F02", "#7570B3", "#66A61E", "grey"), 
                        breaks = c("ABC transporter", "AMG", "Lysogeny", "PHROG", "CDS")) +
                        # values = c("yellow", "#D95F02", "orange", "#7570B3", "#E7298A", "#66A61E", "grey"), 
                        # breaks = c("ABC transporter", "AMG", "Lysis", "Lysogeny", "Peptidase", "Structural/Replication", "CDS")) +
      theme_genes() +
      theme(legend.position = "none",
            text = element_text(size = 10)) +
      guides(fill = guide_legend(nrow = 1)) +
      scale_x_continuous(labels = scales::comma, limits = c(0, 400)) + 
      labs(y = "",
           x = "Nucleotide position (kb)",
           fill = "")
  
  A <- ggwrap(plot, 4)
  
# plot genome map for DNA_vOTU_6575
  plot <-
    ggplot(filter(df, NAME == "DNA_vOTU_6575"), aes(xmin = start_position, xmax = end_position, y = NAME, forward = strandedness, fill = FUNCTION)) +
      geom_gene_arrow() +
      scale_fill_manual(values = c("yellow", "#D95F02", "#7570B3", "#66A61E", "grey"), 
                        breaks = c("ABC transporter", "AMG", "Lysogeny", "PHROG", "CDS")) +
      theme_genes() +
      theme(legend.position = "none",
            text = element_text(size = 10)) +
      scale_x_continuous(labels = scales::comma, limits = c(0, 100)) + 
      labs(y = "",
           x = "Nucleotide position (kb)",
           fill = "")
  
  B <- plot
  
# plot genome map for DNA_vOTU_2761
  plot <-
    ggplot(filter(df, NAME == "DNA_vOTU_2761"), aes(xmin = start_position, xmax = end_position, y = NAME, forward = strandedness, fill = FUNCTION)) +
      geom_gene_arrow() +
      scale_fill_manual(values = c("yellow", "#D95F02", "#7570B3", "#66A61E", "grey"), 
                        breaks = c("ABC transporter", "AMG", "Lysogeny", "PHROG", "CDS")) +
      theme_genes() +
      theme(legend.position = "none",
            text = element_text(size = 10)) +
      scale_x_continuous(labels = scales::comma, limits = c(0, 200)) + 
      labs(y = "",
           x = "Nucleotide position (kb)",
           fill = "") 

  C <- ggwrap(plot, 2)
    
# plot genome map for DNA_vOTU_5919
  plot <-
    ggplot(filter(df, NAME == "DNA_vOTU_5919"), aes(xmin = start_position, xmax = end_position, y = NAME, forward = strandedness, fill = FUNCTION)) +
      geom_gene_arrow() +
      scale_fill_manual(values = c("yellow", "#D95F02", "#7570B3", "#66A61E", "grey"), 
                        breaks = c("ABC transporter", "AMG", "Lysogeny", "PHROG", "CDS")) +
      theme_genes() +
      theme(legend.position = "bottom",
            text = element_text(size = 10)) +
      scale_x_continuous(labels = scales::comma, limits = c(0, 400)) + 
      labs(y = "",
           x = "Nucleotide position (kb)",
           fill = "") 
  
  legend <- get_legend(plot)
  plot <- plot + theme(legend.position = "none")
  
  D <- ggwrap(plot, 4)
  
# plot genome map for DNA_vOTU_707
  plot <-
    ggplot(filter(df, NAME == "DNA_vOTU_707"), aes(xmin = start_position, xmax = end_position, y = NAME, forward = strandedness, fill = FUNCTION)) +
      geom_gene_arrow() +
      scale_fill_manual(values = c("yellow", "#D95F02", "#7570B3", "#66A61E", "grey"), 
                        breaks = c("ABC transporter", "AMG", "Lysogeny", "PHROG", "CDS")) +
      theme_genes() +
      theme(legend.position = "none",
            text = element_text(size = 10)) +
      scale_x_continuous(labels = scales::comma, limits = c(0, 200)) + 
      labs(y = "",
           x = "Nucleotide position (kb)",
           fill = "") 
  
  E <- ggwrap(plot, 2)
  
  plot <- plot_grid(A, B, C, D, E, legend, ncol = 1, rel_heights = c(3, 1.3, 2, 3, 2, 0.5))
  
  ggsave(plot, width = 170, height = 225, units = "mm", dpi = 300,
     file = here("Figures/Figure S8.pdf"), device = cairo_pdf)
    
```

## Discussion

# Supplementary Figure S9 - Rank abundance of A jumbo phages and B CAZyme-carrying vOTUs

```{r Supplementary Figure S9 - Rank abundance of jumbo phages}

# create a data frame of rank abundance of jumbo phages
  df <- 
    DNA_vOTU_reads %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    select(NAME, LENGTH, PHAGE, contains("Sage")) %>%
    pivot_longer(contains("Sage"), names_to = "SAMPLE", values_to = "RPKB") %>% 
    group_by(SAMPLE) %>% 
    mutate(RANK = rank(-RPKB)) %>% 
    mutate(QUARTILE = (RANK/10196)*100) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))
  
# plot the rank abundance of jumbo phages
  plot <- 
    ggplot(filter(df, LENGTH > 200000), aes(x = DEPTH, y = QUARTILE)) +
      geom_boxplot(aes(group = DEPTH)) +
      facet_grid(cols = vars(SITE)) +
      scale_y_continuous(limits = c(0, 100),
                         breaks = seq(0, 100, 10)) +
      scale_x_continuous(limits = c(10, 125),
                           breaks = seq(20, 120, 20)) +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)",
           y = "Rank abundance (%)",
           title = "")
  
  A <- plot_grid(plot, labels = "A", label_size = 15)
  
# create a data frame of the rank abundance of CAZyme-carrying vOTUs
  df <- 
    DNA_vOTU_reads %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    select(NAME, LENGTH, PHAGE, contains("Sage")) %>%
    pivot_longer(contains("Sage"), names_to = "SAMPLE", values_to = "RPKB") %>% 
    group_by(SAMPLE) %>% 
    mutate(RANK = rank(-RPKB)) %>% 
    mutate(QUARTILE = (RANK/10196)*100) %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                               grepl("100", SAMPLE) ~ 100,
                               grepl("80", SAMPLE) ~ 80,
                               grepl("60", SAMPLE) ~ 60,
                               grepl("40", SAMPLE) ~ 40,
                               grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak"))
  
# plot the rank abundance of CAZyme-carrying vOTUs
  plot <-
    ggplot(filter(df, NAME %in% filter(AMG_annotations, !cazy_hits == "")$NAME), aes(x = DEPTH, y = QUARTILE)) +
      geom_boxplot(aes(group = DEPTH)) +
      facet_grid(cols = vars(SITE)) +
      scale_y_continuous(limits = c(0, 100),
                         breaks = seq(0, 100, 10)) +
      scale_x_continuous(limits = c(10, 125),
                           breaks = seq(20, 120, 20)) +
      theme(text = element_text(size = 10)) +
      labs(x = "Soil depth (cm)",
           y = "Rank abundance (%)",
           title = "")
  
  B <- plot_grid(plot, labels = "B", label_size = 15)
  
  plot <- plot_grid(A, B, ncol = 1, rel_heights = c(1, 1))
  
  ggsave(plot, width = 170, height = 200, units = "mm", dpi = 300,
         file = here("Figures/Figure S9.pdf"), device = cairo_pdf)

```

# Supplementary Figure S10 - Relative abundance of hosts of viruses carrying auxiliary metabolic genes

```{r Supplementary Figure S10 - Relative abundance of hosts of viruses carrying auxiliary metabolic genes}

# create a data frame of relative abundance of hosts of viruses carrying AMGs
  df <- 
    DNA_vOTU_reads %>% 
    filter(NAME %in% AMG_annotations$NAME) %>% 
    mutate(across(contains("Sage"), ~./(LENGTH/1000))) %>%
    mutate(across(contains("Sage"), ~(./sum(.)))) %>%
    # mutate(HOST_CLASS = case_when(is.na(HOST_CLASS) ~ "Unknown",
    #                                TRUE ~ HOST_CLASS)) %>%
    group_by(HOST_CLASS) %>% 
    summarise(across(contains("Sage"), sum)) %>% 
    pivot_longer(-HOST_CLASS, names_to = "SAMPLE", values_to = "PROP") %>% 
    group_by(HOST_CLASS) %>% 
    mutate(HOST_CLASS = case_when(sum(PROP) < 0.1 ~ "Other",
                                   TRUE ~ HOST_CLASS)) %>% 
    group_by(HOST_CLASS, SAMPLE) %>% 
    summarise(PROP = sum(PROP), .groups = "drop") %>%
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ 115,
                             grepl("100", SAMPLE) ~ 100,
                             grepl("80", SAMPLE) ~ 80,
                             grepl("60", SAMPLE) ~ 60,
                             grepl("40", SAMPLE) ~ 40,
                             grepl("20", SAMPLE) ~ 20)) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) %>% 
    group_by(HOST_CLASS, DEPTH, SITE) %>% 
    summarise(PROP = mean(PROP), .groups = "drop") %>% 
    mutate(HOST_CLASS = factor(HOST_CLASS, levels = c("Other", "BetaPseudomonadota", "Vicinamibacteria", "Bacteroidia", "GammaPseudomonadota", "Clostridia", "DeltaPseudomonadota", "Actinomycetia"))) # %>% 
  # filter(!is.na(HOST_CLASS))

# plot the relative abundance of hosts of viruses carrying AMGs 
  plot <-
    ggplot(df, aes(x = DEPTH, y = PROP, fill = HOST_CLASS)) +
      geom_area(colour = "white", position = "stack", outline.type = "upper", linewidth = 0.3) +
      facet_grid(cols = vars(SITE)) +
      scale_x_continuous(limits = c(20, 120),
                         breaks = seq(20, 120, 20)) +
      scale_y_continuous(limits = c(0, 1.01),
                         breaks = seq(0, 1, 0.25)) +
      scale_fill_manual(breaks = c("Actinomycetia", "DeltaPseudomonadota", "Clostridia", "GammaPseudomonadota", "Bacteroidia", "Vicinamibacteria", "BetaPseudomonadota", "Other"),
                        labels = c("*Actinomycetia*", "*DeltaPseudomonadota*", "*Clostridia*", "*GammaPseudomonadota*", "*Bacteroidia*", "*Vicinamibacteria*", "*BetaPseudomonadota*", "Other"),
                        values = c(brewer.pal(7, "Dark2"), "grey80")) +
      theme(text = element_text(size = 10), 
            legend.text = element_markdown()) +
      guides(fill = guide_legend(reverse = T,
                                 override.aes = list(linewidth = 0.5, colour = c(rep("white", 8))))) +
      labs(x = "Soil depth (cm)",
           y = "Mean proportional abundance",
           fill = "")
  
  ggsave(plot, width = 170, height = 100, units = "mm", dpi = 300,
         file = here("Figures/Figure S10.pdf"), device = cairo_pdf)

```

# Correlation of microbial macro diversity and viral micro diversity

```{r Correlation of microbial macro diversity and viral micro diversity}
  
# create a data frame of microbial macro diversity
  microbial_diversity <-
    OTU_coverage %>%
    mutate(across(contains("Sage"), ~round(.))) %>%
    select(NAME, contains("Sage")) %>%
    column_to_rownames("NAME") %>%
    otu_table(., taxa_are_rows = T) %>%
    estimate_richness(., measures = c("Shannon")) %>%
    as_tibble(., rownames = "SAMPLE")

# join viral diversity to microbial macro diversity
  df <- 
    inner_join(micro_pi, microbial_diversity, by = "SAMPLE") %>% 
    mutate(DEPTH = case_when(grepl("113", SAMPLE) | grepl("115", SAMPLE) ~ "115 cm",
                             grepl("100", SAMPLE) ~ "100 cm",
                             grepl("80", SAMPLE) ~ "80 cm",
                             grepl("60", SAMPLE) ~ "60 cm",
                             grepl("40", SAMPLE) ~ "40 cm",
                             grepl("20", SAMPLE) ~ "20 cm")) %>% 
    mutate(SITE = case_when(grepl("Sage1", SAMPLE) ~ "Hilly grassland",
                            grepl("Sage2", SAMPLE) ~ "Garry Oak")) 

# compute correlation coefficients of viral diversity and microbial macro diversity
  stat_correlation <- 
    df %>%
    nest(data = everything()) %>% 
    mutate(cor_test = map(data, ~cor.test(.x$PI_MEAN, .x$Shannon, method = "pearson")),
           lm_test = map(data, ~lm(.x$PI_MEAN ~ .x$Shannon)),
           cor_tidied = map(cor_test, tidy),
           lm_tidied = map(lm_test, tidy)) %>%
    unnest(c(cor_tidied, lm_tidied), names_sep = ".") %>% 
    filter(!lm_tidied.term == "(Intercept)") %>% 
    mutate(label = paste("r = ", signif(cor_tidied.estimate,3), 
                         "\nslope = ", signif(lm_tidied.estimate, 3),
                         "\np = ", signif(cor_tidied.p.value, 3))) %>% 
    select(label) %>% 
    mutate(XPOS = NA,
           YPOS = NA)
  
# plot the correlation of microbial macrco diversity and viral micro diversity
  plot <-
    ggplot(df, aes(x = Shannon, y = PI_MEAN)) +
      geom_point(aes(shape = SITE, fill = DEPTH), 
                 size = 3, alpha = 0.8, stroke = 0) +
      scale_shape_manual(breaks = c("Hilly grassland", "Garry Oak"),
                         labels = c("Hilly grassland", "Garry Oak"),
                         values = c(22, 24)) +
      scale_fill_manual(breaks = c("20 cm", "40 cm", "60 cm", "80 cm", "100 cm", "115 cm"), 
                        values = browns_palette) +
      scale_y_continuous(limits = c(0, 0.0005),
                         breaks = seq(0, 0.0005, 0.0001),
                         labels = scales::comma) +
      scale_x_continuous(limits = c(5.8, 6.6),
                         breaks = seq(5.8, 6.6, 0.2)) +
      guides(fill = guide_legend(override.aes = list(shape = 21)),
             shape = guide_legend(override.aes = list(fill = "white", stroke = 1, size = 3))) +
      theme(legend.position = "right",
            text = element_text(size = 10)) +
      labs(x = "Microbial macro diversity (Shannon's H)",
           y = "Viral micro diversity (pi)",
           fill = "Soil depth:", 
           shape = "Site:")
  
  # ggsave(plot, width = 170, height = 100, units = "mm", dpi = 300,
  #        file = here("Figures/Figure S12.pdf"), device = cairo_pdf)

```
